### 2021-12-10 paper/paper_I_comparison.Rmd
added a style reference, a csl, and an example .rmd paper

---
title: "paper_I_methods_comparison"
author: "William Hemstrom"
date: "11/21/2021"
output:
  word_document:
    reference_docx: "style_ref.docx"
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    fig_width: 7
    fig_height: 6
bibliography: "citations.bib"
csl: "council-of-science-editors-author-date.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespacing
  - \usepackage{sectsty}\sectionfont{\fontsize{12}{12}\selectfont}
  - \usepackage{sectsty}\subsectionfont{\normalfont\itshape\fontsize{12}{12}\selectfont}
  - \usepackage[round]{natbib}
indent: true
sansfont: Times New Roman
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2);library(GeneArchEst); library(RefManageR); library(snpR)
bib <- ReadBib("citations.bib")
```

# Abstract:

# Introduction:

<!-- - Can put some of the background info into the results as an additional results paragraph. -->

Given that the rapid rate of global, human induced environmental change is driving an ongoing "sixth mass extinction" [@Gerardo2021], there is a dire need for conservationists to estimate how selective pressures caused by environmental change are likely to impact the species they study over the coming decades. The ability of a population to persist in the face of environmental change depends, essentially, on two factors: the rate of environmental change and genetics of the traits under selection [@Lynch1993]. An understanding of the genetic structure of phenotypic variation is therefore a critical component of any attempt to forecast demographic outcomes for populations in changing environments [@Ellegren2008; @Wilson2010].

While methodologies to incorporate genetic data into demographic models have existed for a while now [see @Burger1995 or @Chevin2010, for example], our massively expanded access high resolution sequencing offers us the opportunity to make much more informed demographic predictions that incorporate far more information about the genetics of a trait under selection. This has led to the development of the *genomic vulnerability* framework, which harnesses genomic data from individuals across varied landscapes to explore how genetic variance correlates with environmental conditions [@Fitzpatrick2015]. This data can then be used in conjunction with climate forecasting to quantify the vulnerability of populations to predicted environmental changes [@Fitzpatrick2015; @Bay2018]. Outlier or association studies, such as Bay et al. [-@Bay2017], expand on this framework by using variation associated with specific environmental variables to directly forecast population demographics.

However, the actual power of these methods to accurately estimate demographic trajectories during adaptation have not been assessed. In order to determine the actual utility of these approaches for genetic risk assessment and demographic forecasting, we need to know how reasonable their predictions are and quantify any bias they tend to include. *Here, we use simulation to generate a known demographic outcome for a population with a known genetic architecture for a trait under selection. We then compare these results to forecasted demographic outcomes produced by using association testing and other approaches to integrate genetic data into demographic models in order to validate the validity of those models.*

# Results:

## Demographic Simulations with a Known Genetic Architecture:

In order to evaluate the power of GP, GWAS, and RF for generating forecasts of demographic outcomes during selective processes, we first simulated demographic outcomes for populations under selection for a trait with several different known genetic architectures. To do so, we first simulated genomic data for 500 individuals using a coalescent approach [@Staab2015], then sampled locus effect sizes off of a BayesB distribution [@Meuwissen2001], derived phenotypes for individuals following the animal model assuming a normally distributed environmental effect ($h^2$ = 0.5), and then simulated survival and growth under constant, directional change in the phenotypic optimum according to Bürger and Lynch [-@Burger1995]. We tracked population size and genetic variance ($\sigma_{g}^{2}$) in each population during each generation of the simulation. We repeated this process 100 times for each architecture in order to explore the range of demographic outcomes in each case. We used this procedure to test four different genetic architectures: one with 30 causal loci/sites, one with 100, one with 300, and one in which all loci had an effect on phenotype (hereafter the BayesA architecture). In each case, we tuned the demographic and selection parameters to ensure that populations went extinct in roughly 150 generations.

We then constructed a 95% prediction interval for time-to-extinction for each architecture. These intervals were slightly wider for the 30 sites data, but were similar overall (Figure X). The overall demographic patterns were also similar: in each case, population size remained near the carrying capacity for the majority of the simulation, then abruptly crashed. $\sigma_{g}^{2}$ declined at a steady rate throughout the simulation, nearing zero at approximately the same time the population went extinct.

## Demographic Forecasting using genomics:

To evaluate the ability of existing methods to predict the demographic outcomes that we observed with our simulated data, we tested two different types of methodologies for predicting the genetic architecture of a trait: 1) causal loci identification or 2) phenotype prediction via the joint distribution of all loci simultaneously. The causal loci identification is typified by Genome Wide Association Studies (GWAS), which are widely used to identify loci associated with phenotypic variance at a specific trait [@Visscher2017]. Over the last 15 or so years, GWAS approaches have been adopted incredibly broadly broadly and are used to answer a huge range of medical, evolutionary, and ecological questions [@Visscher2017; @Visscher2012]. Today, GWAS is arguably the standard approach to causal loci identification.

In contrast, prediction via the joint distribution of all loci is typically done via Genomic Prediction (GP) or Random Forest (RF) approaches. These methods jointly consider all loci, even those with very little or no effect, in order to maximize phenotypic prediction accuracy for a given trait [@DelosCampos2013; @Meuwissen2001]. GP models have been used primarily in the agricultural and animal science fields, but have also seen some use in an ecological context [CITE]. RF, a Machine Learning method which has been increasingly used for ecological genetics and association studies [@Lunetta2004; @Molinaro2011; @Bay2018; @Goldstein2011; @Meng2009], approaches prediction similarly by considering the joint effect of all sequenced loci, although it by bypasses the model construction step by using algorithmically derived models without direct human involvement in order to maximize prediction accuracy.

To generate demographic forecasts from GWAS candidate loci identification, we ran a GWAS using each set of phenotypes and selected all loci for $p \leq 0.05$ in each case. We then used just these phenotypes to construct a basic multiple linear regression against phenotype. We then ran the demographic simulations as before, but used the animal model to predicting phenotypes each generation, where $G$ was predicted for each individual by prediction off of this regression.

We found that this approach was mostly unable to generate realistic estimates of time-to-extinction. In each test architecture, GWAS predicted a significantly more rapid time-to-extinction than observed in the control data. This problem became more severe as the number of causal loci increased, with overlap in the time-to-extinction prediction interval _only_ with 30 causal loci (Figure X). The demographic patterns and changes in $\sigma_{g}^{2}$ were quite similar to those in the control data, however, with a constant population size followed by an abrupt crash and a constant decline, respectively. Interestingly, we observed several distinctly different demographic trajectories in both the 30 sites and 100 sites outcomes with were observed repeatedly across the different simulations.

For these GP and RF, we generated demographic forecasts from GP and RF by predicting $G$ for each individual using the respective model in each generation. Like GWAS, neither GP nor RF approach was able to generate reasonable estimates for time-to-extinction: in each case and for each test architecture, time-to-extinction estimates were much lower than in the control data (Figure X). Critically, these approaches both predicted an immediate, dramatic drop in $\sigma_{g}^{2}$ in all test architectures (Figure X). We observed this decline even if chromosomes were randomly shuffled between individuals without recombination in the absence of selection.

As an alternative approach, we also drew BayesB hyperparameters [@Meuwissen2001] from the posterior distribution produced by GP and then sampled phenotypes as with the control data. Since a single hyperparameter draw can produce a wide range of actual architectures due to placing effects on loci with different minor allele frequencies, we filtered the resulting phenotypic distributions by taking only those that were similar to the control phenotypic distribution. The prediction intervals for time-to-extinction from this approach partially overlapped the control data for all architectures other than the BayesA architecture, but still consistently underestimated the true time-to-extinction (Figure X). $\sigma_{g}^{2}$ estimates were followed a more similar tragectory to the control data, but were highly variable (Figure X).

## Demographic Forecasting using Quantitative Genetics:

In addition to the genomic methods above, we also conducted demographic forecasting using a quantitative genetics model which relies only on the heritability of a trait, not on genomic data or any kind of genetic architecture estimate [@Burger1995]. This model (hereafter BL) predicted population persistence for all architectures other than 300 sites, for which the time-to-extinction was considerably underestimated (Figure X).

# Discussion:

## Demographic Forecasting with Genomics:

Prior to conducting this study, we hypothesized that a GWAS candidate-loci based approach to genetic architecture estimation would result in consistent underestimates of time-to-extinction that worsen as the true number of effect loci increase. We reasoned that this would occur due to GWAS’s difficulty with detecting rare causal alleles, those with small effect sizes, or both [@Maher2008; @Nolte2017; @Visscher2017]. As a result, GWAS studies should typically underestimate the number of causal loci and overestimate the average effect sizes of those it does detect. Since populations with few loci of strong effect should have strong initial responses to selection but less longevity as variance is exhausted as environmental change continues [@Kardos2021], GWAS should as a result underestimate time-to-extinction.

This is exactly what our simulations demonstrated: under the GWAS model, there are fewer loci with effects, each of which has a much larger effect on average than in the real data (Figure X). As a result, population sizes do not initially change in size at all, since the strong response in the underlying mean additive genetic component of phenotype is able to easily match pace with the rate of environmental change (X). This was followed in a sharp decline $\sigma_{g}^{2}$ in the population declines rapidly (Figure X), which is accompanied shortly thereafter by a precipitous decline in population size (Figure X). Interestingly, we observed that GWAS predicted several distinctly different possible demographic outcomes when the number of causal loci was low (30 or 100 sites) that mirrored similar trends in the control data. In each case, these patterns was caused by the stochastic loss of major effect loci due to genetic drift during the early period of the simulation, which resulted in a reduced ability to respond as selection pressures increased.

In contrast, we initially expected GP or RF approaches to overestimate time-to-extinction, which very clearly did not happen. These approaches both jointly consider all loci in order to predict individual BVs. Due to linkage, we expected that this would result in the attribution of effect to too many loci, and the spreading out of the size of the effect amongst these loci, resulting in an overestimation of the number of effective loci and an underestimation of the size of their average genomic effects. This would result in a slower initial response to selection, but a more sustained response over time and thus a longer average time-to-extinction [@Kardos2021]. While GP did indeed predict more loci of smaller effect (Figure X), both GP and RF predicted a slow initial response, but then an extremely quick decline in population size (Figure X). This rapid decline was caused in both cases by an immediate and pronounced decline in genetic variance (Figure X).

Unlike declines in the other models, the variance declines for GP and RF were not caused by selection or drift, since reshuffling chromosomes between individuals without replacement caused a similar decline in estimated variance. While the reason for this is unclear, it likely relates to the spreading of effect estimates across linked loci, even those that are not physically linked. Breaking these linkages up by reshuffling may cause a loss in genetic variance by spreading effects away from the effect “peaks” present in the initial estimates and thus causing all subsequent individuals to have more similar genomic compositions. The same issue may explain the decline of variance in RF derived estimates as well. Regardless, it is clear that estimating the effects (or importance) of individual effects in GP and RF result in less accurate demographic estimates. Unsurprisingly, then, drawing architectures from the hyperparameter posteriors of a GP estimation did produce marginally better estimates without the sharp decline in genetic variance, although the posterior very clearly overestimated the number of effective loci and underestimated their average effect size (Figure X).

## Demographic Forecasting with Classical Quantitative Genetics:

Given the poor performance of the GWAS, GP, and RF approaches to demographic forecasting, we also attempted to use a quantitative genetic approach to forecast demographic outcomes without success. Bürger and Lynch's [-@Burger1995] method did not predict extinction during the 250 generations that we simulated for most architectures. This is not a particularly surprising result given the structure of the BL model and of quantitative genetics models in general.

Most quantitative genetics models are based on the “animal” model, which offers a straightforward way to incorporate genetic information into evolutionary and ecological models by modeling individual phenotypes (P) as the sum of the genetic (G) and environmental factors (E) which influence that trait [@Falconer1996]: 
$$\begin{eqnarray} P = G + E \end{eqnarray}$$ 
The breeder’s equation is perhaps the most natural way to apply the animal model to predict phenotypic responses to selection:
$$ \begin{eqnarray} R &=& h^{2}S \\ h^{2} &=& \frac{\sigma_{g}^{2}}{\sigma_{p}^{2}} \end{eqnarray}$$
where the per-generation response to selection ($R$) is a simple product of the strength of selection ($S$) and the ratio between the additive, or directly heritable, genetic ($\sigma_{g}^{2}$) and phenotypic ($\sigma_{p}^{2}$) variance [together narrow-sense heritability, or $h^{2}$, see @Lush1943]. However, this equation does not always perform well in wild populations [@Morrissey2010].

While the animal model and its direct derivations have seen extensive use in both agricultural and ecological settings to model selection, both artificial and natural [@Germain2016; @Kostick2021; @Krist2021; @Pettersen2018; @Zhao2015], more complex derivations are potentially more useful for modeling population responses to selection specifically. The BL model sits in this category alongside a set of others which use $h^2$ and a set of biological and demographic parameters to predict population persistence during environmental change [@Burger1995, @Chevin2010; @Gomulkiewicz2009; @Lynch1993; @Osmond2017]. These models [specifically the BL model and @Lynch1993's very similar model] have been recently implemented to estimate the critical rates of environmental change for hatching dates in *Parus major* [@Gienapp2013], and thermal stress in *Drosophila birchii* [@Willi2009].

However, as we have shown here, these models may not always perform well. They have two core issues: first, they only include genetic information in the form of a $h^2$ estimate. Given that the details of genomic architecture should produce markedly different biological outcomes, even when heritability is identical [@Chevalet1994; @Walsh2003; @Walsh2018], they may produce misleading estimates for some populations. In our case, however, the problem is likely that these models assume an infinitesimal model of genetic variation, under which selection does not change the underlying genetic variation for a trait [@Fisher1919; @Robertson1960], which is unlikely to be a reasonable approximation unless a trait is incredibly polygenic.

They therefore allow for the indefinite maintenance of genetic variance ($\sigma_{g}^{2}$) during selection, since the removal of alleles due to selection does not result in a reduction of $\sigma_{g}^{2}$ if there are infinite possible alleles that contribute to a given trait [@Lynch1993; @Burger1995]. A population's core ability to respond to selection therefore stay consistent over time, and populations go extinct only due to stochastic processes or if the constant lag between the population's mean phenotype and the population optimum is too large [@Burger1995; @Lynch1993]. They cannot simply run out of genetic variation as they hit physiological limits or reach a maximum possible phenotype determined by the initial standing genetic variation.

In contrast, in a real-world population where the number of causal loci for every trait is finite, $\sigma_{g}^{2}$ will decline as selection purges unfit alleles unless mutation can generate new variance apace. In our simulations, we see this clearly: $\sigma_{g}^{2}$ declines over time due to selection for our observed results and for our models which incorporate more genetic information. It is therefore not surprising that the BL model overestimated the capacity of our population to respond to selection over successive generations.

Unfortunately, our results are likely generalizable, since the infinitesimal model is likely to be unreasonable for most populations of conservation concern. For the model to be a close approximation, mutation (or other processes that introduce genetic variation) must resupply enough adaptive genetic variation in order to maintain a population's adaptive capacity. Specifically, this means that a mutation needs to supply causal variants which help extend a phenotype beyond the previous maximum or minimum as selection fixes variants that previously contributed to one tail of the phenotypic distribution. For populations experiencing rapid changes to their environment, as is often the case populations of concern, this may simply not be possible, and thus the infinitesimal model is likely to underestimate demographic impacts over time, as it did here.

## Conclusion:

Ultimately, since the infinitesimal model is not realistic for virtually any trait in any population, classical quantitative genetics approaches to demographic forecasting are not likely to produce reasonable estimations of demographic outcomes for populations of conservation concern. To remedy this, more genetic information must be incorporated into demographic models if reasonably accurate predictions are desired. However, it is clear that neither the causal loci focused GWAS approach or joint-site phenotype prediction focused GP and RF approaches produce reasonable forecasts, since the former approach will tend to miss the potentially important contributions of loci will small effects to a population's adaptive capacity and the latter have some intrinsic issues which cause them to erroneously predict a drastic decline in genetic variance in subsequent generations. New methods, such as True Genomic Architecture Computation (TGAC, see Hemstrom et al. 2022), may help sidestep the problems with these approaches.

# Methods:

## Genotype and Phenotype Simulation:

We produced simulated genotype data for 500 whole genome datasets using the program scrm [@Staab2015] assuming ten 10mb long chromosomes, a effective population size ($N_{e}$) of 1000, a recombination rate ($\rho$) of $2x10^{-8}$ per base, and a mutation rate ($\mu$) of $1x10^{-8}$ per base. We then simulated two sets of phenotypes by applying modified BayesB distributions, where the parameter $\pi$, which denotes the probability that a locus has no effect on phenotype was substituted for the parameter *sites*, which denotes the exact number of loci with an effect. By making this change, we ensured that we knew exactly how many sites had an effect on phenotype for each of our simulations. Phenotypes *P* for each individual *i* were determined using the animal model summing across each *j* allele:
$$ \begin{eqnarray} P_{i} &=& G_{i} + E_{i} \\ G_{i} &=& \sum_{i=1}^{n}\sum_{j=1}^{2} p_{i,j}a_{j} \\ E_{i} &\sim& N(0,\sigma_{e}) \\ \sigma_{e} &=& \sqrt{\frac{\sigma_{g}^{2}}{h^{2}} - \sigma_{g}^{2}} \end{eqnarray}$$
where *G* is the additive genetic component of phenotype, *E* is the environmental component, $p_{i,j}$ is the count of the allele *j* in individual *i*, $a_{j}$ is the effect of allele *j*,  $\sigma_{e}^{2}$ is the variance of the environmental effects, $\sigma_{g}^{2}$ is the variance of the genetic effects, and $h^{2}$ is the narrow-sense heritability.

We created three sets of phenotypes with $sites =$ 30, 100, or 300, $s =$ 1, $h^{2} =$ 0.5, and $df =$ 5. We simulated one additional set of phenotypes using a BayesA distribution, which like, BayesB, draws effects from a scaled *t*-distribution with scale *s*, but assumes that all loci have an effect [equivalent to a *sites* value equal to the number of segregating sites in the genome, see @Meuwissen2001].

## Demographic Simulations with a Known Architecture:

We simulated demographic responses to selection with all four test architectures. We used a demographic model similar to that proposed by  Bürger & Lynch [-@Burger1995]. In this model, the optimum phenotype $\theta$ increases each generation *t* linearly by *k*, but varies stochastically each generation following a normal distribution with a standard deviation of $\sigma_{\theta}$:
\begin{eqnarray} &\sigma_{t} = \sigma_{t-1} + k + x_{t} \\ &x_{t} \sim N(0, \sigma_{\theta}^{2})\end{eqnarray}
The probability that individual *i* survives generation *t* ($s_{i,t}$) is determined by their phenotype $P_{i}$, the selection optimum in that generation $\theta_{t}$, and the strength of selection $\omega$:
$$ \begin{eqnarray} s_{i,t} = e^{-\frac{(p_{i} - \theta_{t})^{2}} {2\omega^{2}}} \end{eqnarray} $$
We incorporated density dependence by randomly culling individuals down to *K* survivors after each generation. 
Surviving individuals were then assigned a sex at random and randomly mated to reproduce to $N\beta$ surviving offspring, where *N* is the number of survivors and $\beta$ is the average number of offspring per surviving individual. We allowed chromosomes to recombine by randomly placing *r* recombination events, where *r* was determined using a Poisson distribution with $\lambda$ = 1, for a mean of one recombination event per chromosome per generation. We chose this value of $\lambda$ to match the value of $\rho$ used during the initial coalescent simulations used to generate the test genomes.

Individual phenotypes were calculated as above in each generation and subjected to selection. We assumed that narrow-sense heritability was equal to broad-sense heritability, and that the scale of the environmental effects on phenotype did not change over time. Heritability, therefore, could change if the amount of population genetic variance changed across generations. If N was ever < 2 or all survivors were of one sex, the population was considered to have gone extinct. We modeled the population without any overlap in generations. We tracked the population size *N* and the population genetic variance $\sigma_{g}^{2}$ in each generation until extinction. In each case, *K* was set to 250 and $\beta$ to 2, such that the populations would return to 500 individuals (the initial number of simulated genomes) if culled to *K* without selection. The other demographic and selection parameters were tuned for each test architecture such that the population went extinct in roughly 150 generations (see Table SX). In order to determine the range of possible demographic outcomes for each architecture, we ran the simulation 100 times for each and calculated 95% prediction intervals for time-to-extinction.

## Forecasting with Genomics:

Since real whole-genome data is essentially never complete, we first a reasonable amount of missing genotype data. We added missing data to our genotypic data by drawing a probability of missing data at each locus from a normal distribution of mean 0.05, sd = 0.2, such that each locus had a different probability of having missing data. Missing data was then added by a binomial draw at each genotype based on the locus-specific missing data probability. We then imputed missing genotypes using the Beagle software package [@Browning2007] without using a reference panel. We repeated this process for each test architecture.

We then conducted a GWAS model using a relatedness-corrected association test using the GMMAT R package [@Chen2016]. To identify candidate loci, selected all SNPs where $p \leq 1x10^{-5}$. We then conducted demographic simulations as above for each of the four test architectures using the same parameters, although the genetic component of each individual phenotype, *G*, was determined by predicting off a multiple linear regression model fit with genotypes at at each of our selected SNPs as the explanatory variables instead of by reference to the known locus effect sizes. In each generation, $\sigma_{g}^{2}$ was calculated using these *G* values. We repeated this procedure 100 times to determine the range of possible predicted demographic outcomes and calculated 95% prediction intervals for time-to-extinction, as before.

For both GP and RF, we first added missing data as in GWAS, then fit the respective model to the observed phenotypes for each of the four test architectures. For GP, we used the BGLR R package [@Perez2014] assuming a BayesB effect size distribution to fit a genomic prediction model using 1,000 burn-in and fit 10,000 iterations. For RF, we used the ranger R package [@Wright2017] with 500 trees and an *mtry* equal to the number of loci. Using more iterations or trees did not seem to improve model fit. In each case, we fit a basic model with phenotype as the response and the genotypes as the explanatory variables without covariates.

We then ran the demographic simulations as before, using the same parameters as the control data. *G* was predicted for each individual in each generation by predicting off of the model using their genotype data. As before, $\sigma_{g}^{2}$ was calculated using these *G* values. We ran each simulation 100 times and calculated 95% prediction intervals as before.

As an additional test, we also conducted demographic forecasting using the posterior BayesB hyper-parameter distribution generated by the genomic prediction model. We then used these parameters to generate allele effect sizes and simulate demographic outcomes as we did with the real data. We repeated this step 1,000 times. However, since a single set of BayesB hyper-parameters can theoretically produce very different architectures if large effects are placed on common vs. rare alleles by chance, we filtered the resulting phenotypic distributions to find those that were reasonable matches by calculating a two-way Kolmogorov-Smirnov statistic comparing the initial control phenotypic distribution (before selection and random mating) to the predicted initial phenotypic distribution [@Conover1998], and took only the top 10% (resulting in 100 predicted demographic outcomes, as for each other approach).

## Demographic Forecasting with Classical Quantitative Genetics:

We also used Bürger and Lynch [@Burger1995]'s methods to forecast demographic outcomes for each of our test architectures. We followed Bürger and Lynch [@Burger1995]'s (denoted BL below) equations to directly iterate across time using the following steps:

1. Use BL equation 7a to determine the expected mean phenotype in this generation ($E[\bar{g}_{t}]$).
2. Use BL equation 7b to determine the variance of the mean phenotype in this generation ($V[\bar{g}_{t}]$).
3. Use $E[\bar{g}_{t}]$ and $V[\bar{g}_{t}]$ alongside BL equations 13, and 9 to determine $\lambda{t}$, the population growth rate for this generation.
4. Determine the number of individuals in the next generation ($N_{t+1}$) using the equation $N_{t+1} = \lambda_{t}N_{t}$, where $N_{t}$ is the number of individuals in this generation. If $N_{t+1}$ was less than two, the population was considered to have gone extinct.
5. If at any point $N_{t} = N_{t+1}$, the demographic and selection pressures were considered not strong enough to cause extinction.

In each case, we allowed the selection intensity to change and parameterized the demographic and selection parameters as we did for the known architecture simulations.

# References:

### 2022-04-08 paper/paper_draft.Rmd
added the paper draft and genotype file

---
title: "MASU Paper"
author: "William Hemstrom"
date: "11/21/2021"
output:
  word_document:
    reference_docx: "style_ref.docx"
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    fig_width: 7
    fig_height: 6
bibliography: "Citations.bib"
csl: "council-of-science-editors-author-date.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespacing
  - \usepackage{sectsty}\sectionfont{\fontsize{12}{12}\selectfont}
  - \usepackage{sectsty}\subsectionfont{\normalfont\itshape\fontsize{12}{12}\selectfont}
  - \usepackage[round]{natbib}
indent: true
sansfont: Times New Roman
fontsize: 12pt
---
```{r, echo=FALSE, include=FALSE}
dat<-read.table("../data/genotypes.geno")                        
meta<- read.table("../data/sample_meta.txt", header=TRUE)
snp.meta<- dat[,1:2]
colnames(snp.meta)<- c("chr", "position")
library(snpR)
dat<- import.snpR.data(dat[,-c(1:2)], snp.meta = snp.meta, sample.meta = meta)
dat<- filter_snps(dat, maf = 0.05, hwe = 0.000001)
number_snps_after_filtering <- nsnps(dat)

#calculating genetic diversity
dat <- calc_ho(dat, c("pop", "phen", "pop.phen"))
get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="ho") $weighted.means
get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
dat <- calc_pi(dat, c("pop"))
get.snpR.stats(dat, c("pop"), stats="pi") $weighted.means
get.snpR.stats(dat, "pop")
dat <- calc_het_hom_ratio(dat, c("pop", "phen", "pop.phen"))
get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
dat <- calc_maf(dat, c("pop"))
get.snpR.stats(dat, c("pop"))
p <- plot_clusters(dat, facets = c("pop.phen"), plot_type = "umap")
dat <- calc_private(dat, c("phen", "pop","pop.phen"))
get.snpR.stats(dat, c("pop"), stats="private") $weighted.means
get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
dat <- calc_pairwise_fst(dat, c("phen", "pop","pop.phen"), boot = 100, boot_par = 2)
get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="fst") $weighted.means
fts <- dat[[2]]
dat <- dat[[1]]
stats <- get.snpR.stats(dat, "pop.phen", type = "pairwise")
head(stats)

gp<- run_genomic_prediction(dat, response = "phen", iterations = 10000, burn_in = 1000, thin = 100)
library(data.table)
cv<- cross_validate_genomic_prediction(dat, response = "phen", iterations = 10000, burn_in = 1000, thin = 100)

saveRDS(dat, "processed_genotypes.RDS")
saveRDS(gp, "gp.RDS")

```
# Methods:
## Genotyping 
We extracted DNA from our samples using the magnetic bead approach [@Ali2016]. We then prepared Restriction Associated Digest (RAD) libraries using the Pst1 restriction enzyme [@Ali2016] and sequenced the results on an Illumina Hi-Seq 4000 using 150bp paired-end sequencing. We aligned the data to the O. mykiss v6 reference genome [@Pearse2019] using Burrows-Wheeler Alignment algorithm Aligner [@Li2009]. We then filtered the data for PCR duplicates and improper pairs using SAMtools [@Li2009]. Next, we called the genotypes by using the SAMtools genotype likelihood models [@Li2009] used in the ANGSD software package with a minimum mapping and base call quality score of 20, a SNP p-value of 1e-8, and a minimum acceptable minor allele frequency of 0.05 [@Korneliussen2014]. We also used a minor allele frequency filter (MAF) and a Hardy-Weinberg equilibrium filter using snpR when calling the genotypes [@Wigginton2005]. 
<!-- flagstat google (data script i wrote) -->

```{r}
mat <- read.table("../angsdput.covMat")
IBS <- data.table::fread("../angsdput.covMat")
nrow(IBS)
IBS <- IBS[-32,-c(97, 32)]
pca_r <- eigen(IBS)
pca <- pca_r$vectors
write.table(pca[,1:20], "pca20.txt", col.names = FALSE, row.names = FALSE, quote=FALSE, sep="\t")
colnames(pca) <- paste0("PC", 1:ncol(pca))
meta <- read.table("../data/sample_meta.txt", header = TRUE)[-32,]
pca <- cbind(pca, meta)

loadings <- pca_r$values/sum(pca_r$values)
loadings <- round(loadings*100, 2)
pca$pop <- as.factor(pca$pop)
pca$phen <- as.factor(pca$phen)

library(ggplot2)
p1 <- ggplot(pca, aes(x = PC1, y = PC2, color = pop, shape = phen)) + theme_bw() +
  scale_color_viridis_d() + geom_point(size = 4) + 
  ggplot2::xlab(paste0("PC1 (", loadings[1], "%)")) + 
  ggplot2::ylab(paste0("PC2 (", loadings[2], "%)"))
p1
pca.centroids <- aggregate(pca[,1:2], list(stream = pca$pop), mean)
combine <- rbind(pca[, 1:2], pca.centroids[,-1])
dcombine<- dist(combine, upper = TRUE) 
dcombine<- as.matrix(dcombine)
dcombine<-dcombine[(nrow(dcombine) - nrow(pca.centroids) + 1):nrow(dcombine), -((nrow(dcombine) - nrow(pca.centroids) + 1):nrow(dcombine))]
dcombine <- t(dcombine) 
dcombine <- as.data.frame(dcombine)
dcombine$pop <- pca$pop
dcombine$phen <- pca$phen
colnames(dcombine)[1:2] <- c("CH", "TK")
# ggplot(dcombine, aes(x=pop, y=d, color=phen)) + geom_boxplot() + theme_bw()
dcombine$phen <- factor(dcombine$phen)
dcombine$pop <- factor(dcombine$pop)
mod <- aov(TK~pop+phen+pop*phen, dcombine)
dcombine$d <- 0
dcombine$d[dcombine$pop=="CH"] <- dcombine$TK[dcombine$pop=="CH"]
dcombine$d[dcombine$pop=="TK"] <- dcombine$CH[dcombine$pop=="TK"]
mod <- aov(d~phen+pop+pop*phen, dcombine[-32,])
mod2 <- aov(d~phen, dcombine[-32,])
anova(mod, mod2)
ggplot(dcombine, aes(x=pop, y=d, color=phen)) + geom_boxplot() + theme_bw() + scale_color_viridis_d(end = 0.5)

# library(ggbiplot)
# wine <- iris [, 1:4]
# wine.pca <- prcomp(wine, center = TRUE, scale. = TRUE)
# df.wine.x <- as.data.frame(wine.pca$x)
# df.wine.x$groups <- iris$Species
# pca.centroids <- aggregate(df.wine.x[,1:4], list(groups = df.wine.x$groups), mean)
# library(ggplot2)
# ggplot(df.wine.x, aes(x=PC1, y=PC2, color=groups)) + geom_point() + theme_bw() + 
#   geom_point(data=pca.centroids, color="red", size=10, alpha=0.5) + facet_wrap(~ groups)
# dist(df.wine.x [,-1])
# combine <- rbind(df.wine.x[,-ncol(df.wine.x)], pca.centroids[,-1])
# dcombine<- dist(combine, upper = TRUE) 
```

##  Diversity Stats
We calculated the observed heterozygosity (Ho), heterozygote/homozygote ratio per individual (Het/Hom), the average number of pairwise differences (pi), and fixation index (Fst) using the snpR R package [@Hohenlohe2010; @Wier1984; @Hemstrom2021]. 
<!-- use table to report diversity stats in results (which ones most diverse) get .snpr stats (in uppoer script)-->
<!-- want ngsadmix plot, manhattan plot, pca chart with centroids (two panel one side pca and box and whiskers plots) -->

```{r}
dat <- run_random_forest(dat, response = "phen", mtry = floor(nrow(dat)/2), num.trees = 100000)
RF <- dat$models
dat <- dat$data
summary(RF$.base_.base)
```

## Association Testing
We performed a basic association test on the data using gmmat to correct for the related among individuals and population structure (). We then performed a random forest analysis in R-studio using the ranger R package [@Wright2017]. We used 100,000 trees and used mtry of half the number of snp's. We used a random forest to predict whether the samples would be resident or migratory. Using the BGLR program, we did a genomic prediction using BayesB distribution to estimate the heritability [@Meuwissen2001]. We used 1000 burn- ins and iterations=10,000. The heritability test was done to determine the degree in which phenotypic variation is due to genetic differences among the salmon populations.

## Population Structure and Migratory Relatedness
We then used NGSadmix (not sure how to describe this). With our data, we also made a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014]. In the PCA, we included centroids of each cluster. 
<!-- Will need to fill in NGSadmix -->

# Results 
```{r}
MASU_data_table <- readr::read_delim("~/MASU_data_table", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE, skip = 4)
MASU_data_table <- as.data.frame(MASU_data_table)
MASU_data_table[,3] <- gsub("%", "", MASU_data_table[,3])
MASU_data_table[,5] <- gsub("%", "", MASU_data_table[,5])
MASU_data_table[,3]<- as.numeric(MASU_data_table[,3])
MASU_data_table[,5]<- as.numeric(MASU_data_table[,5])
mean(MASU_data_table[,3], na.rm = TRUE)
mean(MASU_data_table[,5], na.rm = TRUE)
```

##Genotyping 
<!-- using linux raw count of # of reads -> how many reads we actually mapped (/home/hemstrow/moths/bams/W/run_read_count.sh) bamlists that are not filtered(.bam) and .sort.flt and import into R and take sum of columns that has the read count run the script twice-->

unfiltered reads = 163200129
```{r setup, include=FALSE, eval=FALSE}
library(readr)
sort_read_count <- read.csv("~/Downloads/sort.read_count.txt", header = FALSE)
View(sort_read_count)
sort_read_count <- sum(sort_read_count$V1)

sort_flt_read_count <- read.csv("~/Downloads/sort.flt.read_count.txt", header = FALSE)
View(sort_flt_read_count)
sort_flt_read_count <- sum(sort_flt_read_count$V1)

```

flt read = `r sort_flt_read_count` 
<!-- flt read = 90449126 -->
unfiltered reads = `r sort_read_count` 
<!-- unfiltered reads = 163200129 -->

<!-- report number of snps we called nsnps (dat) and number of loci in IBS data (both in R) report number after filtering -->
number of snps = `r number_snps_after_filtering`
Number of snps= 71175?
number of IBS loci = `r nrow(IBS)`

##  Diversity Stats
<!-- see table X (embedding figures) -->
<!-- talk about any trends like this creek had higher diversity than this  -->
<!-- still need to work on mahanttan plot -> don't think we found anythign significant; there was not a position that greatly differed depending on the sample (in linux) -->
<!-- What are the thresholds at which the Ho values are significant. -->
The mean Ho values are relatively similar between the anadromous and residents individuals and between the CH and TK populations.  
<!-- Having trouble getting the fst values -> but should match up with PCA right?  -->
The fst values between the two populations and between anadromous and resident individuals are low, which indicated that there is not much genetic diversity between the two populations and two phenotypes. 
<!-- report stats in excel table  -->
There are also no private alleles between the two populations, which also indicated low genetic diveristy between the two populations of salmon.

## Association 
<!--phenotypes and genetic basis -->
how to write up the results of the random forest? -> did not find anything significant because the actual results did not match the predicted. report r2 value

## Influence of Anadromy on Population Structure
The distance from the other population's centroid is closer to the anadromous than the resident in one population. This indicates that the anadromous individuals within one population are more closely related to the individuals in the other population. 
<!-- Is there anything else I should add here?  -->
<!-- control shift c --> 
[@Gowen2014; @Ali2016]
Gowen et al.[-@Gowen2014]

### 2022-04-20 paper/paper_draft.Rmd
updated paper draft

@@
 ---
+
+#Introduction 
+MASU salmon have two different types of lifestyles which include either being anadromous or resident. 
 ```{r, echo=FALSE, include=FALSE}
@@
 get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
-dat <- calc_pi(dat, c("pop"))
+dat <- calc_fis(dat, c("pop", "phen"))
+dat <- calc_pi(dat, c("pop", "phen"))
 get.snpR.stats(dat, c("pop"), stats="pi") $weighted.means
-get.snpR.stats(dat, "pop")
-dat <- calc_het_hom_ratio(dat, c("pop", "phen", "pop.phen"))
-get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
-dat <- calc_maf(dat, c("pop"))
-get.snpR.stats(dat, c("pop"))
-p <- plot_clusters(dat, facets = c("pop.phen"), plot_type = "umap")
 dat <- calc_private(dat, c("phen", "pop","pop.phen"))
@@
 get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="fst") $weighted.means
-fts <- dat[[2]]
-dat <- dat[[1]]
-stats <- get.snpR.stats(dat, "pop.phen", type = "pairwise")
-head(stats)
+get.snpR.stats(dat, c("pop", "phen"), stats= c("ho", "pi", "fst")) $weighted.means
 
@@
 colnames(dcombine)[1:2] <- c("CH", "TK")
-# ggplot(dcombine, aes(x=pop, y=d, color=phen)) + geom_boxplot() + theme_bw()
 dcombine$phen <- factor(dcombine$phen)
 dcombine$pop <- factor(dcombine$pop)
-mod <- aov(TK~pop+phen+pop*phen, dcombine)
 dcombine$d <- 0
@@
 dcombine$d[dcombine$pop=="TK"] <- dcombine$CH[dcombine$pop=="TK"]
-mod <- aov(d~phen+pop+pop*phen, dcombine[-32,])
-mod2 <- aov(d~phen, dcombine[-32,])
-anova(mod, mod2)
+mod2 <- lm(d~phen, dcombine[-32,])
 ggplot(dcombine, aes(x=pop, y=d, color=phen)) + geom_boxplot() + theme_bw() + scale_color_viridis_d(end = 0.5)
 
-# library(ggbiplot)
-# wine <- iris [, 1:4]
-# wine.pca <- prcomp(wine, center = TRUE, scale. = TRUE)
-# df.wine.x <- as.data.frame(wine.pca$x)
-# df.wine.x$groups <- iris$Species
-# pca.centroids <- aggregate(df.wine.x[,1:4], list(groups = df.wine.x$groups), mean)
-# library(ggplot2)
-# ggplot(df.wine.x, aes(x=PC1, y=PC2, color=groups)) + geom_point() + theme_bw() + 
-#   geom_point(data=pca.centroids, color="red", size=10, alpha=0.5) + facet_wrap(~ groups)
-# dist(df.wine.x [,-1])
-# combine <- rbind(df.wine.x[,-ncol(df.wine.x)], pca.centroids[,-1])
-# dcombine<- dist(combine, upper = TRUE) 
 ```
@@
 We calculated the observed heterozygosity (Ho), heterozygote/homozygote ratio per individual (Het/Hom), the average number of pairwise differences (pi), and fixation index (Fst) using the snpR R package [@Hohenlohe2010; @Wier1984; @Hemstrom2021].
-<!-- use table to report diversity stats in results (which ones most diverse) get .snpr stats (in uppoer script)-->
-<!-- want ngsadmix plot, manhattan plot, pca chart with centroids (two panel one side pca and box and whiskers plots) -->
 
@@
 summary(RF$.base_.base)
+
+association <- data.table::fread("../association_out_cov.lrt0.gz")
+association$p <- pchisq(association$LRT, 1, lower.tail = FALSE)
+plot_manhattan(assocation, "p", chr = "Chromosome", bp= "Position")
+
 ```
@@
 With our data, we also made a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014].
-In the PCA, we included centroids of each cluster.
+In the PCA, we included the centroids of each cluster.
+The centroid of each cluster is the mean of that cluster of points.
 <!-- Will need to fill in NGSadmix -->
@@
 ##Genotyping 
-<!-- using linux raw count of # of reads -> how many reads we actually mapped (/home/hemstrow/moths/bams/W/run_read_count.sh) bamlists that are not filtered(.bam) and .sort.flt and import into R and take sum of columns that has the read count run the script twice-->
-
-unfiltered reads = 163200129
+The number of unfiltered reads are `r sort_read_count`.
+After filtering, the number of reads were `r sort_flt_read_count`.
+We called a total of `r number_snps_after_filtering` snps and the number of IBS loci were `r nrow(IBS)`.
 ```{r setup, include=FALSE, eval=FALSE}
@@
 <!-- unfiltered reads = 163200129 -->
-
-<!-- report number of snps we called nsnps (dat) and number of loci in IBS data (both in R) report number after filtering -->
 number of snps = `r number_snps_after_filtering`
@@
 ##  Diversity Stats
-<!-- see table X (embedding figures) -->
-<!-- talk about any trends like this creek had higher diversity than this  -->
-<!-- still need to work on mahanttan plot -> don't think we found anythign significant; there was not a position that greatly differed depending on the sample (in linux) -->
-<!-- What are the thresholds at which the Ho values are significant. -->
 The mean Ho values are relatively similar between the anadromous and residents individuals and between the CH and TK populations.
-<!-- Having trouble getting the fst values -> but should match up with PCA right?  -->
 The fst values between the two populations and between anadromous and resident individuals are low, which indicated that there is not much genetic diversity between the two populations and two phenotypes.
-<!-- report stats in excel table  -->
-There are also no private alleles between the two populations, which also indicated low genetic diveristy between the two populations of salmon.
+There are also no private alleles between the two populations, which also indicated low genetic diversity between the two populations of salmon.
 
 ## Association 
-<!--phenotypes and genetic basis -->
-how to write up the results of the random forest? -> did not find anything significant because the actual results did not match the predicted. report r2 value
+The random forest predicted 25 out of 52 anadromous individuals correctly and 17 out of 44 resident individuals correctly for a total prediction accuracy of 38.6%.
 
 ## Influence of Anadromy on Population Structure
-The distance from the other population's centroid is closer to the anadromous than the resident in one population.
+We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.
+We found that the andadromous individuals clustered signficiantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419).
 This indicates that the anadromous individuals within one population are more closely related to the individuals in the other population.
-<!-- Is there anything else I should add here?  -->
+<!-- The p value (phenR) is 0.03981775. The p value for the phenotypes is 0.0419.  -->
+
 <!-- control shift c --> 

### 2022-04-20 paper/paper_draft.Rmd
moved NGS admix files, added ngsadmix methods to the paper

@@
 ## Population Structure and Migratory Relatedness
-We then used NGSadmix (not sure how to describe this).
-With our data, we also made a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014].
-In the PCA, we included the centroids of each cluster.
-The centroid of each cluster is the mean of that cluster of points.
+In order to detect any underlying population structuring within our stream systems, we used the software package NGSadmix to group samples into *k* clusters for each value of *k* between one and seven [@Skotte693].
+To do so, we first called genotype posterior probailites using ANGSD with the same parameters above in addition to (fill in here).
+To assess variance in clustering within a given *k* value, we ran NGSadmix ten times for each value of *k* and combined the resulting data using CLUMPP [@Jakobsson2007].
+We used the pophelper and snpR packages to handle files and produce plots [@Hemstrom2021; @Francis2017].
+
+For comparison, we also conducted a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014].
+To look for patterns in relatedness between resident and andromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample the mean for the other population.
+
+<!-- include that we did a one-way anova to determine if either phenotype was more closely related to the mean of the opposite population -->
+
+```{r ngsadmix,include=FALSE}
+setwd("../ngsadmix/")
+ngsadmix <- plot_structure(".qopt", paste0(meta$pop, "_", meta$phen), facet.order = c("CH_A", "CH_R", "TK_A", "TK_R"), k = 1:7)
+ngsadmix <- ngsadmix$plot + xlab("Population/Phenotype")
+```
+
 <!-- Will need to fill in NGSadmix -->

### 2022-04-29 paper/paper_draft.Rmd
Updated MASU paper with rough draft of introduction

@@
 #Introduction
-MASU salmon have two different types of lifestyles which include either being anadromous or resident. 
+
+<!-- Declines in migratory fishes -->
+<!-- 	Trends and potential causes. -->
+
+<!-- Importance of Masu -->
+<!-- 	Vulnerability -->
+<!-- 		Very prone to fishing pressure -->
+<!-- 		Very limited distribution -->
+
+<!-- Life History -->
+<!-- 	What decides if an indivividual is a resident/anadromous? -->
+
+<!-- Genetics of resident vs anadromous -->
+
+<!-- What did we do and find? -->
+<!-- 	Thesis -->
+Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in the recent years.
+For example, on the Pacific coast there has been significant decreases in diversity within salmonid species due hydrologic and dam alterations.
+Atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams.
+Overall, the diversity within migratory fish species have largely declined due to these dams blocking the fishes' migratory paths.
+One species of migratory fish that we will be discussing more in depth is the MASU salmon.
+
+MASU salmon, or the cherry salmon, have the most limited distribution among the Pacific salmon.
+Additionally, since they spend a lot of their life in freshwater rivers or streams, this makes them more inclined to be exposed to fishing and sportfishing pressures.
+Since they are more vulnerable than other salmon species, it is important to look into the details that influence their life cycles of being either resident or anadromous.
+Anadromous salmon are initially born in the rivers then migrate to the seas where they develop larger body sizes.
+These salmon eventually migrate back to the rivers in order to breed.
+The resident salmon in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives (Morita and Nagasawa)
+
+MASU salmon and their life cycles have been known to be primarily influenced by environmental factors such as temperature and other factors such as the salmon's body size, growth rate and lipid stores.
+Environmental factors such as climate change have been known to impact the decision of the salmon to be either resident or anadromous.
+It has been shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased.
+The increase in resident males due to increasing temperatures have led to a significant decrease in the number of anadromous individuals.
+Instead of looking into environmental influences, which have already beeen heavily studied, in this paper, we will be examining if there is a genetic basis within MASU salmon that will allow us to determine the salmon's life cycle.
+
+For comparison, within the O.
+mykiss species or rainbow trout, which is also a species of fish that exhibit either resident or anadromous life cycles, there is strong evidence suggesting a genetic basis on life history.
+The first evidence of a genetic basis for O.
+mykiss was when Neave (1944) conducted common garden and transplant experiments with sympatric rainbow trout and steelhead.
+After marking and releasing the progeny of the rainbow trout and steelhead, he found that life cycles were heritable after a larger proportion of rainbow trout offspring were recaptured over many years.
+Additionally, crosses between anadromous O.
+mykiss individuals resulted in the most smolts while crosses between resident individuals produced the fewest number of smolts.
+In a recent study by Hess et al.
+(2016), univariate analyses and a multivariate random forest (RF) machine learning algorithm was used to perform the association mapping of of 15 239 single nucleotide (SNPs) for adult migration-timing phenotype in steelhead.
+Using a univariate linear model resulted in the finding of three SNPs from a gene that was strongly linked with migration timing.
+Using multivariate analyses with RF resulted in the finding of 15 additional SNPs that were connected to 60% of the trait variation.
+
+Like how the rainbow trout and steelhead have a clear genetic basis for their life histories, we also wanted to investigate the genetic influences on MASU salmon and their life cycles using similar methods.
+We performed a random forest on the data, did association tests, and found diversity statistics, which did show any clear indicator of the genetic difference between the anadromous and resident MASU salmon.
+However, when creating a Principle Component Analysis chart (PCA) to look for patterns in relatedness between resident and anadromous individuals, we found that the anadromous individuals clustered significantly closer to the cluster center for individuals from the other population.
+Overall, from the DNA analysis methods used in this paper, there is not strong evidence that there is a genetic basis for MASU salmon life histories, but further studies using a bigger sample size would need to be conducted in order to investigate the influences of genetics on MASU salmon and their life cycles.
+
 ```{r, echo=FALSE, include=FALSE}
@@
 We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.
-We found that the andadromous individuals clustered signficiantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419).
+We found that the anadromous individuals clustered signficiantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419).
 This indicates that the anadromous individuals within one population are more closely related to the individuals in the other population.

### 2022-05-18 paper/paper_draft.Rmd
MASU paper with updated discussion section

@@
 <!-- 	Thesis -->
+
 Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in the recent years.
@@
 
+<!-- Anadromous salmon -> dams causing them to not be able to jump over dams -> is we explore the genetic influences of anadromy and residents in MASU salmon then we can tell where to focus our conservation efforts.
+-->
+
+<!-- Parr size was generally larger and maturation was more common at more southern latitudes -->
+
+<!-- Population density and temperature two most important factors affecting the growth of the fish.  -->
+
+<!-- superior development promotes residency (early maturation) -->
+
+<!-- need to do citations -->
 ```{r, echo=FALSE, include=FALSE}
@@
 
-<!-- Will need to fill in NGSadmix -->
-
 # Results 
@@
 
+# Discussion 
+## Results Discussion
+According to the PCA, there is overlap between the anadromous and resident individuals from both populations.
+In other words, there are not clear distinct clusters that separate the anadromous individuals from the resident individuals.
+This indicates there is not strong evidence for a genetic influence that determines the MASU salmon's life cycle.
+
+## Ramifications for MASU salmon
+In these populations it is possible that residency and anadromy within MASU salmon are not solely genetically controlled and that we must take into consideration the environmental influences.
+When comparing MASU salmon to steelhead, steelhead and their life cycles are more heavily influenced by genetic factors.
+As of now, MASU salmon and their life cycles are mainly influenced my environmental factors, but further studies with larger sample sizes would need to be conducted in order to see if any genetic influences play a role.
+In other words, we would need still need to figure out if MASU salmon life cycles are influenced by the additive effects of genetics and environment, the multiplying effetcs of genetics and environment, or simply a result of purely environmental effects.
+
+## Ramifications for Salmonids in General 
+In terms of the importance of this work, figuring out if MASU salmon life cycles have genetics influences will go towards their conservation efforts.
+For example, within steelhead fish, anadromous individuals will have to migrate upstream in order to move towards the ocean.
+On some occurrences the anadromous steelhead will be stopped midway towards their journey to the ocean due to manmade dams.
+This has caused a reduction in diversity within these fish by causing an apparent increase in resident individuals within the population.
+Knowing whether a fish is anadromous or resident would be helpful because people can advocate for dams to built in areas where there are a lot of anadromous fish.
+This same logic can also be applied towards the conservation of MASU salmon.
+
 <!-- control shift c --> 

### 2022-05-20 paper/paper_draft.Rmd
added processed RF and gp, changed paper chunks to read and not run things

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, eval = FALSE, include = FALSE)
+library(ggplot2); library(RefManageR); library(snpR)
+
+```
 
@@
 <!-- need to do citations -->
-```{r, echo=FALSE, include=FALSE}
+```{r, include=FALSE, eval=TRUE}
 dat<-read.table("../data/genotypes.geno")                        
@@
 library(snpR)
-dat<- import.snpR.data(dat[,-c(1:2)], snp.meta = snp.meta, sample.meta = meta)
-dat<- filter_snps(dat, maf = 0.05, hwe = 0.000001)
-number_snps_after_filtering <- nsnps(dat)
+# dat<- import.snpR.data(dat[,-c(1:2)], snp.meta = snp.meta, sample.meta = meta)
+# dat<- filter_snps(dat, maf = 0.05, hwe = 0.000001)
+# number_snps_after_filtering <- nsnps(dat)
+# 
+# #calculating genetic diversity
+# dat <- calc_ho(dat, c("pop", "phen", "pop.phen"))
+# get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="ho") $weighted.means
+# get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
+# dat <- calc_fis(dat, c("pop", "phen"))
+# dat <- calc_pi(dat, c("pop", "phen"))
+# get.snpR.stats(dat, c("pop"), stats="pi") $weighted.means
+# dat <- calc_private(dat, c("phen", "pop","pop.phen"))
+# get.snpR.stats(dat, c("pop"), stats="private") $weighted.means
+# get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
+# dat <- calc_pairwise_fst(dat, c("phen", "pop","pop.phen"), boot = 100, boot_par = 2)
+# get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="fst") $weighted.means
+# get.snpR.stats(dat, c("pop", "phen"), stats= c("ho", "pi", "fst")) $weighted.means
+# 
+# gp<- run_genomic_prediction(dat, response = "phen", iterations = 10000, burn_in = 1000, thin = 100)
+# library(data.table)
+# cv<- cross_validate_genomic_prediction(dat, response = "phen", iterations = 10000, burn_in = 1000, thin = 100)
 
-#calculating genetic diversity
-dat <- calc_ho(dat, c("pop", "phen", "pop.phen"))
-get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="ho") $weighted.means
-get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
-dat <- calc_fis(dat, c("pop", "phen"))
-dat <- calc_pi(dat, c("pop", "phen"))
-get.snpR.stats(dat, c("pop"), stats="pi") $weighted.means
-dat <- calc_private(dat, c("phen", "pop","pop.phen"))
-get.snpR.stats(dat, c("pop"), stats="private") $weighted.means
-get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
-dat <- calc_pairwise_fst(dat, c("phen", "pop","pop.phen"), boot = 100, boot_par = 2)
-get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="fst") $weighted.means
-get.snpR.stats(dat, c("pop", "phen"), stats= c("ho", "pi", "fst")) $weighted.means
+# saveRDS(dat, "processed_genotypes.RDS")
+# saveRDS(gp, "gp.RDS")
 
-gp<- run_genomic_prediction(dat, response = "phen", iterations = 10000, burn_in = 1000, thin = 100)
-library(data.table)
-cv<- cross_validate_genomic_prediction(dat, response = "phen", iterations = 10000, burn_in = 1000, thin = 100)
-
-saveRDS(dat, "processed_genotypes.RDS")
-saveRDS(gp, "gp.RDS")
+dat <- readRDS("../processed_genotypes.RDS")
+gp <- readRDS("gp.RDS")
 
@@
 
-```{r}
+```{r, eval=TRUE, include=FALSE}
 mat <- read.table("../angsdput.covMat")
@@
 
-```{r}
-dat <- run_random_forest(dat, response = "phen", mtry = floor(nrow(dat)/2), num.trees = 100000)
-RF <- dat$models
-dat <- dat$data
-summary(RF$.base_.base)
+```{r, eval=TRUE, include=FALSE}
+# dat <- run_random_forest(dat, response = "phen", mtry = floor(nrow(dat)/2), num.trees = 100000)
+# RF <- dat$models
+# dat <- dat$data
+# summary(RF$.base_.base)
+# saveRDS("RF", "RF.RDS")
+RF <- readRDS("RF.RDS")
 
@@
 association$p <- pchisq(association$LRT, 1, lower.tail = FALSE)
-plot_manhattan(assocation, "p", chr = "Chromosome", bp= "Position")
+plot_manhattan(association, "p", chr = "Chromosome", bp= "Position", log.p = TRUE)
 
@@
 
-```{r ngsadmix,include=FALSE}
+```{r ngsadmix, include=FALSE}
 setwd("../ngsadmix/")
@@
 # Results 
-```{r}
+```{r, include=FALSE}
 MASU_data_table <- readr::read_delim("~/MASU_data_table", 
@@
 We called a total of `r number_snps_after_filtering` snps and the number of IBS loci were `r nrow(IBS)`.
-```{r setup, include=FALSE, eval=FALSE}
+```{r read, include=FALSE}
 library(readr)

### 2022-05-27 paper/paper_draft.Rmd
updated discussion section

--- 2022-05-18 paper/paper_draft.Rmd
@@
 According to the PCA, there is overlap between the anadromous and resident individuals from both populations.
-In other words, there are not clear distinct clusters that separate the anadromous individuals from the resident individuals.
-This indicates there is not strong evidence for a genetic influence that determines the MASU salmon's life cycle.
+Although having overlapping clusters does not exclude the possibility of a genetic influence, our findings did not clearly show associated genetic elements between resident and anadromous fish.
+As mentioned in the previously to look for patterns in relatedness between resident and andromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample and the mean for the other population.
+It is important to note that we did find that anadromous individuals were more closely related to individuals in the other population.
+<!-- This can potentially be attributed to where the samples were collected.
+For example, if the anadromous individuals from one stream were collected in an area that was geographically near the individuals in the other stream, this could explain this observation.
+(not sure how to interpret the metadata to confirm this) -->
+We also made a manhattan plot and did not find many individuals that had a locus with a high p-value.
+This indicates that there is not an area in the genome that is strongly associated with the given trait (anadromy vs.
+resident).
+The random forest also did not have strong predictive power for anadromy and residency because it could not accurately group the individuals to their actual life cycle.
+Overall there is not strong evidence for a genetic influence that determines the MASU salmon's life cycle.
 
 ## Ramifications for MASU salmon
-In these populations it is possible that residency and anadromy within MASU salmon are not solely genetically controlled and that we must take into consideration the environmental influences.
+In these populations it is possible that residency and anadromy within MASU salmon are not solely genetica lly controlled and that we must take into consideration the environmental influences.
 When comparing MASU salmon to steelhead, steelhead and their life cycles are more heavily influenced by genetic factors.
@@
 
+<!-- talk about genetic mechanism in steelhead and how that varies in environmental  -->
 ## Ramifications for Salmonids in General 

### 2022-05-30 paper/paper_draft.Rmd
updated introduction and added an abstract

@@
 
+#Abstract
+Within Masu salmon (Oncorhynchus masou), individuals can exhibit one of two life cycles: resident or anadromous.
+The expresson of either life history is mainly influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced (CITE).
+Here, we explore if there are genetic factors that influence the development of either residency or anadromy.
+In order to look for patterns of relatedness among resident and anadromous individuals, we made a PCA chart, did association tests, and made a random forest (RF).
+Ultimately, we did not find any significant associated genetic elements, but further research with a larger sample size is needed in order to fully assess potential genetic influences on anadromy and residency.
+
 #Introduction
@@
 
-Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in the recent years.
-For example, on the Pacific coast there has been significant decreases in diversity within salmonid species due hydrologic and dam alterations.
-Atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams.
-Overall, the diversity within migratory fish species have largely declined due to these dams blocking the fishes' migratory paths.
-One species of migratory fish that we will be discussing more in depth is the MASU salmon.
+Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in the recent years (CITE).
+In North America, for example,there have been significant decreases indiversity within pacific salmonid species due hydrologic and dam alterations (CITE), and atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams (CITE).
+Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to these dams blocking the fishes' migratory paths (CITE- need a research paper for this number)
 
-MASU salmon, or the cherry salmon, have the most limited distribution among the Pacific salmon.
-Additionally, since they spend a lot of their life in freshwater rivers or streams, this makes them more inclined to be exposed to fishing and sportfishing pressures.
-Since they are more vulnerable than other salmon species, it is important to look into the details that influence their life cycles of being either resident or anadromous.
-Anadromous salmon are initially born in the rivers then migrate to the seas where they develop larger body sizes.
-These salmon eventually migrate back to the rivers in order to breed.
-The resident salmon in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives (Morita and Nagasawa)
+Masu salmon (Oncorhynchus masou, otherwise known as cherry salmon) have the most limited distribution among the Pacific salmon (CITE).
+Since they spend a greater portion of their life in freshwater rivers or streams than most other migratory salmon, they are particularly more exposed to fishing and sportfishing pressures (CITE).Their increased vulnerability makes it critical to understand the details that influence and control migratory life history variance in the species.
+Masu in particular (like Oncorhynchus mykiss, steelhead salmon/rainbow trout) are partially migratory, with large resident and anadromous populations (CITE).
+As is typical, anadromous masu are initially born in the rivers then migrate to the seas where they develop larger body sizes, then eventually migrate back to their natal rivers in order to breed (CITE).
+Resident masu in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives (Morita and Nagasawa)
 
-MASU salmon and their life cycles have been known to be primarily influenced by environmental factors such as temperature and other factors such as the salmon's body size, growth rate and lipid stores.
-Environmental factors such as climate change have been known to impact the decision of the salmon to be either resident or anadromous.
-It has been shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased.
-The increase in resident males due to increasing temperatures have led to a significant decrease in the number of anadromous individuals.
-Instead of looking into environmental influences, which have already beeen heavily studied, in this paper, we will be examining if there is a genetic basis within MASU salmon that will allow us to determine the salmon's life cycle.
+The decision to migrate in masu salmon is thought to be primarily influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced (CITE).
+Consequentially, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous (CITE).
+For example, X et al (YEAR) has shown that the number of resident males have increased over the last X years while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased, which has led to a significant decrease in the number of anadromous individuals overall.
 
-For comparison, within the O.
-mykiss species or rainbow trout, which is also a species of fish that exhibit either resident or anadromous life cycles, there is strong evidence suggesting a genetic basis on life history.
-The first evidence of a genetic basis for O.
-mykiss was when Neave (1944) conducted common garden and transplant experiments with sympatric rainbow trout and steelhead.
-After marking and releasing the progeny of the rainbow trout and steelhead, he found that life cycles were heritable after a larger proportion of rainbow trout offspring were recaptured over many years.
-Additionally, crosses between anadromous O.
-mykiss individuals resulted in the most smolts while crosses between resident individuals produced the fewest number of smolts.
-In a recent study by Hess et al.
-(2016), univariate analyses and a multivariate random forest (RF) machine learning algorithm was used to perform the association mapping of of 15 239 single nucleotide (SNPs) for adult migration-timing phenotype in steelhead.
-Using a univariate linear model resulted in the finding of three SNPs from a gene that was strongly linked with migration timing.
-Using multivariate analyses with RF resulted in the finding of 15 additional SNPs that were connected to 60% of the trait variation.
+In contrast, the genetic factors that contribute to migration in masu salmon have not been well studied.
+Despite the prevalence of environmental control of migration in masu, it is not unreasonable to expect genetic elements to play some role in governing migratory decisions in the species.
+For example, within steelhead, there is strong evidence suggesting both an environmental and genetic basis for migratory life history.
+In 1944, Naeve conducted common garden and transplant experiments with sympatric rainbow trout and steelhead and found that migratory status was heritable after observing that rainbow trout offspring were considerably more likely to be resident themselves, whereas  crosses between anadromous individuals resulted in the most anadromous smolts.
+It is not surprising, therefore, that a recent study by Hess et al.
+(2016), on 15,239 single nucleotide polymorphisms (SNPs) found of three SNPs from a gene that was strongly linked with adult migration timing, and that 18 total SNPs that were connected to 60% of the trait variation.
+Despite this, however, growth rates are known to strongly influence anadromy in the species—individuals which grow faster as juveniles are far more likely to stay as residents than otherwise (CITE).
+Indeed, migration in the species seems to be controlled jointly the interplay of genetic and environmental factors:
 
-Like how the rainbow trout and steelhead have a clear genetic basis for their life histories, we also wanted to investigate the genetic influences on MASU salmon and their life cycles using similar methods.
-We performed a random forest on the data, did association tests, and found diversity statistics, which did show any clear indicator of the genetic difference between the anadromous and resident MASU salmon.
-However, when creating a Principle Component Analysis chart (PCA) to look for patterns in relatedness between resident and anadromous individuals, we found that the anadromous individuals clustered significantly closer to the cluster center for individuals from the other population.
-Overall, from the DNA analysis methods used in this paper, there is not strong evidence that there is a genetic basis for MASU salmon life histories, but further studies using a bigger sample size would need to be conducted in order to investigate the influences of genetics on MASU salmon and their life cycles.
+The degree of genetic control of migration in masu, however, is still not well known.
+To improve our understanding of the mechanisms behind residency and anadromy in the species,  we sequenced a set of anadromous and resident masu individuals using high-resolution Restriction Associated Digest (RAD) sequencing.
+However, we did not find any clear indicator of the genetic difference between ouranadromous and resident Masu salmon.
+While this may suggest that there is not that there is a strong genetic basis for Masu salmon migratory life history variation, further studies using a bigger sample size from a larger geographic range would need to be conducted in order to fully investigate the influences of genetics on Masu salmon migratory life cycles.
 

### 2022-05-31 paper/paper_draft.Rmd
Updated discussion

@@
 # Discussion
-## Results Discussion
-According to the PCA, there is overlap between the anadromous and resident individuals from both populations.
-Although having overlapping clusters does not exclude the possibility of a genetic influence, our findings did not clearly show associated genetic elements between resident and anadromous fish.
-As mentioned in the previously to look for patterns in relatedness between resident and andromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample and the mean for the other population.
-It is important to note that we did find that anadromous individuals were more closely related to individuals in the other population.
-<!-- This can potentially be attributed to where the samples were collected.
-For example, if the anadromous individuals from one stream were collected in an area that was geographically near the individuals in the other stream, this could explain this observation.
-(not sure how to interpret the metadata to confirm this) -->
-We also made a manhattan plot and did not find many individuals that had a locus with a high p-value.
-This indicates that there is not an area in the genome that is strongly associated with the given trait (anadromy vs.
-resident).
-The random forest also did not have strong predictive power for anadromy and residency because it could not accurately group the individuals to their actual life cycle.
-Overall there is not strong evidence for a genetic influence that determines the MASU salmon's life cycle.
+## Genetic Associations with Migration
 
-## Ramifications for MASU salmon
-In these populations it is possible that residency and anadromy within MASU salmon are not solely genetica lly controlled and that we must take into consideration the environmental influences.
-When comparing MASU salmon to steelhead, steelhead and their life cycles are more heavily influenced by genetic factors.
-As of now, MASU salmon and their life cycles are mainly influenced my environmental factors, but further studies with larger sample sizes would need to be conducted in order to see if any genetic influences play a role.
-In other words, we would need still need to figure out if MASU salmon life cycles are influenced by the additive effects of genetics and environment, the multiplying effetcs of genetics and environment, or simply a result of purely environmental effects.
+We did not find any significant genetic associations with migratory phenotype (see Figure X).
+This indicates that in the streams we examined, there may not be any loci in the genome that strongly contribute to migratory life history decisions.
+It is not therefore surprising that the random forest which we conducted also did not have strong predictive power for anadromy and residency, since it would not be able to accurately group the individuals to their actual life history if there are no genetic elements that contribute to migratory phenotype.
 
-<!-- talk about genetic mechanism in steelhead and how that varies in environmental  -->
-## Ramifications for Salmonids in General 
-In terms of the importance of this work, figuring out if MASU salmon life cycles have genetics influences will go towards their conservation efforts.
-For example, within steelhead fish, anadromous individuals will have to migrate upstream in order to move towards the ocean.
-On some occurrences the anadromous steelhead will be stopped midway towards their journey to the ocean due to manmade dams.
-This has caused a reduction in diversity within these fish by causing an apparent increase in resident individuals within the population.
-Knowing whether a fish is anadromous or resident would be helpful because people can advocate for dams to built in areas where there are a lot of anadromous fish.
-This same logic can also be applied towards the conservation of MASU salmon.
+However, our study included samples from only a small region at the center of the masu range.
+It is possible that if we sampled from other areas, we could uncover loci that influence migratory life history phenotypes under environmental conditions that differ from those in our study systems.
+This could occur if the effect of migratory genes depends on environmental conditions (GxE).
+This phenomena is seen in steelhead, where variance in environmental conditions across the species’ range can determine whether life histories are more genetically or environmentally based (CITE).
+
+## Patterns of Relatedness Between Anadromous and Resident Individuals
+
+We found that although our two populations (TK and CH) were genetically distinct from one another, there was substantial overlap between the anadromous and resident individuals within each population, indicating that resident and anadromous individuals freely interbreed within populations.
+Interestingly, we did find that anadromous individuals were significantly more closely related to individuals from the stream from which they did not originate.
+This could potentially be attributed to where the samples were collected.
+For example, if the anadromous individuals from one stream were collected closer to the river mouth than residents, and thus geographically closer to the other stream, this could explain why we observed that they were more genetically similar to individuals from that stream.
+Alternatively, if anadromous individuals are more likely to stray than residents (since they may return to the wrong river system), and their offspring are more likely to be anadromous themselves, then this could also produce the pattern we observed.
+
+However, this hypothesis could only be true if there are transgenerational environmental or genetic correlations with anadromy.
+Given that we did not find any obvious genetic associations with anadromy, the latter seems more likely.
+For example, transgenerational environmental correlations anadromy could exist if anadromous individuals tend to breed in areas where there are less nutrients available, since smaller body sizes induce the decision of an anadromous life cycle.
+Under these circumstances, the offspring of migrants would be more likely to be migrants themselves, allowing for the localized build up of alleles brought in by strays.
+
+## Conservation Ramifications
+Our results suggest that, at least in some areas, residency and anadromy within masu salmon are not significantly genetically controlled and that fisheries managers and conservation groups should therefore focus more on the possible environmental correlations with migration (such as temperature and resource availability) when planning conservation efforts targeting masu salmon.
+Unfortunately, this also means that genetic monitoring for the anadromous potential of a stream is probably often not possible, since it is not possible to predict migratory phenotype from genotypic data.
+These results may not extrapolate across the entire masu range: further studies with larger sample sizes from more areas could help determine if migratory life history decisions are influenced by the additive or multiplicative effects of genetics and environment or simply a result of purely environmental effects across the full range of the species.
 

### 2022-06-01 paper/paper_draft.Rmd
wording changes to methods and results

@@
 ## Genotyping 
-We extracted DNA from our samples using the magnetic bead approach [@Ali2016].
-We then prepared Restriction Associated Digest (RAD) libraries using the Pst1 restriction enzyme [@Ali2016] and sequenced the results on an Illumina Hi-Seq 4000 using 150bp paired-end sequencing.
-We aligned the data to the O.
-mykiss v6 reference genome [@Pearse2019] using Burrows-Wheeler Alignment algorithm Aligner [@Li2009].
+We extracted DNA from our samples using the magnetic beads and prepared Restriction Associated Digest (RAD) libraries using the Pst1 restriction enzyme according to Ali et al [-@Ali2016], then sequenced the results on an Illumina Hi-Seq 4000 using 150bp paired-end sequencing.
+We aligned the data to the *O.
+mykiss* v6 reference genome [@Pearse2019] using Burrows-Wheeler Alignment algorithm [@Li2009].
 We then filtered the data for PCR duplicates and improper pairs using SAMtools [@Li2009].
-Next, we called the genotypes by using the SAMtools genotype likelihood models [@Li2009] used in the ANGSD software package with a minimum mapping and base call quality score of 20, a SNP p-value of 1e-8, and a minimum acceptable minor allele frequency of 0.05 [@Korneliussen2014].
-We also used a minor allele frequency filter (MAF) and a Hardy-Weinberg equilibrium filter using snpR when calling the genotypes [@Wigginton2005].
+Next, we called  genotypes by using the SAMtools genotype likelihood models [@Li2009] used in the ANGSD software package with a minimum mapping and base call quality score of 20, a SNP p-value of 1e-8, and a minimum acceptable minor allele frequency of 0.05 [@Korneliussen2014].
+We then removed loci out of Hardy-Weinberg Equilibrium ($p \le 0.000001$) according to Wigginton et al.'s [@Wigginton2005] method using the snpR R package [@Hemstrom2021].
 <!-- flagstat google (data script i wrote) -->
@@
 ##  Diversity Stats
-We calculated the observed heterozygosity (Ho), heterozygote/homozygote ratio per individual (Het/Hom), the average number of pairwise differences (pi), and fixation index (Fst) using the snpR R package [@Hohenlohe2010; @Wier1984; @Hemstrom2021].
+We calculated the observed heterozygosity ($H_{o}$), the average number of pairwise differences ($\pi$), number of private alleles, and fixation index ($F_{st}$) between populations using the snpR R package [@Hohenlohe2010; @Wier1984; @Hemstrom2021].
 
@@
 ## Association Testing
-We performed a basic association test on the data using gmmat to correct for the related among individuals and population structure ().
-We then performed a random forest analysis in R-studio using the ranger R package [@Wright2017].
-We used 100,000 trees and used mtry of half the number of snp's.
-We used a random forest to predict whether the samples would be resident or migratory.
-Using the BGLR program, we did a genomic prediction using BayesB distribution to estimate the heritability [@Meuwissen2001].
-We used 1000 burn- ins and iterations=10,000.
+We performed a basic association test against migratory phenotype using ANGSD using the first 20 Principal Component Analysis components as covariates to correct for the related among individuals and population structure [@Korneliussen2014].
+To increase our association resolution, we used used genotype likelihoods instead of called gneotypes to conduct our association test, generated in ANGSD using the same parameters described above.
+In order to see if any more complex genetic relationships could predict migratory phenotype, we then performed a random forest analysis in R-studio using the ranger R package [@Wright2017].
+We used 100,000 trees and used mtry of half the number of called SNPs.
+
+<!-- We also used the BGLR program to perform a genomic prediction using BayesB distribution to estimate the heritability [@Meuwissen2001].
+We used 1,000 burn-ins and 10,000 estimation iterations.
 The heritability test was done to determine the degree in which phenotypic variation is due to genetic differences among the salmon populations.
+-->
 
 ## Population Structure and Migratory Relatedness
-In order to detect any underlying population structuring within our stream systems, we used the software package NGSadmix to group samples into *k* clusters for each value of *k* between one and seven [@Skotte693].
-To do so, we first called genotype posterior probailites using ANGSD with the same parameters above in addition to (fill in here).
+In order to detect any underlying population structuring within our stream systems, we used the software package NGSadmix to group samples into *k* clusters for each value of *k* between one and seven [@Skotte693] using the genotype likelihoods mentioned above.
 To assess variance in clustering within a given *k* value, we ran NGSadmix ten times for each value of *k* and combined the resulting data using CLUMPP [@Jakobsson2007].
-We used the pophelper and snpR packages to handle files and produce plots [@Hemstrom2021; @Francis2017].
+We used the pophelper and snpR packages to handle files and produce plots, respectively [@Hemstrom2021; @Francis2017].
 
@@
 To look for patterns in relatedness between resident and andromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample the mean for the other population.
+We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.
 
@@
 After filtering, the number of reads were `r sort_flt_read_count`.
-We called a total of `r number_snps_after_filtering` snps and the number of IBS loci were `r nrow(IBS)`.
+We called a total of `r number_snps_after_filtering` SNPs and `r nrow(IBS)` IBS loci.
 ```{r read, include=FALSE}
@@
 
-flt read = `r sort_flt_read_count` 
+<!-- flt read = `r sort_flt_read_count`  -->
 <!-- flt read = 90449126 -->
-unfiltered reads = `r sort_read_count` 
+<!-- unfiltered reads = `r sort_read_count`  -->
 <!-- unfiltered reads = 163200129 -->
-number of snps = `r number_snps_after_filtering`
-Number of snps= 71175?
-number of IBS loci = `r nrow(IBS)`
+<!-- number of snps = `r number_snps_after_filtering` -->
+<!-- Number of snps= 71175? -->
+<!-- number of IBS loci = `r nrow(IBS)` -->
 
 ##  Diversity Stats
-The mean Ho values are relatively similar between the anadromous and residents individuals and between the CH and TK populations.
-The fst values between the two populations and between anadromous and resident individuals are low, which indicated that there is not much genetic diversity between the two populations and two phenotypes.
-There are also no private alleles between the two populations, which also indicated low genetic diversity between the two populations of salmon.
+The mean $H_{o}$ values are relatively similar between the anadromous and residents individuals and between the CH and TK populations.
+The $F_{ST}$ values between the two populations and between anadromous and resident individuals were very low, which indicated that there is not much genetic diversity between the two populations and two phenotypes.
+There are also no private alleles between the two populations, which also indicated low genetic differentiation between the two populations of salmon.
 
 ## Association
-The random forest predicted 25 out of 52 anadromous individuals correctly and 17 out of 44 resident individuals correctly for a total prediction accuracy of 38.6%.
+We did not find any strong associations with migratory life history status at any loci, and the random forest only successfully predicted 25 out of 52 anadromous individuals correctly and 17 out of 44 resident individuals correctly for a total prediction accuracy of 38.6%.
 
 ## Influence of Anadromy on Population Structure
-We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.
-We found that the anadromous individuals clustered signficiantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419).
-This indicates that the anadromous individuals within one population are more closely related to the individuals in the other population.
+ We found that the anadromous individuals clustered significantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419).
+This indicates that the anadromous individuals within one population were more closely related to the individuals in the other population than were residents.
 <!-- The p value (phenR) is 0.03981775. The p value for the phenotypes is 0.0419.  -->

### 2022-06-01 paper/paper_draft.Rmd
updated intro

--- 2022-05-31 paper/paper_draft.Rmd
@@
 ---
-title: "MASU Paper"
+title: "No Detectable Evidence of Genetic Associations Between Anadromous and Resident Masu Salmon Sampled within Hokkaido"
 author: "William Hemstrom"
@@
 #Abstract
-Within Masu salmon (Oncorhynchus masou), individuals can exhibit one of two life cycles: resident or anadromous.
-The expresson of either life history is mainly influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced (CITE).
-Here, we explore if there are genetic factors that influence the development of either residency or anadromy.
-In order to look for patterns of relatedness among resident and anadromous individuals, we made a PCA chart, did association tests, and made a random forest (RF).
-Ultimately, we did not find any significant associated genetic elements, but further research with a larger sample size is needed in order to fully assess potential genetic influences on anadromy and residency.
 
-#Introduction
+Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in recent years (Pringle et al., 2000).
+In North America, for example,there have been significant decreases in diversity within pacific salmonid species due to hydrologic and dam alterations, and atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams (Pringle et al., 2000).
+Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to these dams blocking the fishes' migratory paths (Deinet et al., 2020).
 
-<!-- Declines in migratory fishes -->
-<!-- 	Trends and potential causes. -->
+Masu salmon (Oncorhynchus masou, otherwise known as cherry salmon) have the most limited distribution among the Pacific salmon.
+Since they spend a greater portion of their life in freshwater rivers or streams than most other migratory salmon, they are particularly exposed to fishing and sportfishing pressures (Miyakoshi et al., 2009).
+Their increased vulnerability makes it critical to understand the details that influence and control migratory life history variance in the species.
+Masu in particular (like steelhead salmon/rainbow trout Oncorhynchus mykiss and other species to a lesser degree) are partially migratory, with large resident and anadromous populations (Morita et al., 2014).
+As is typical, anadromous masu are initially born in the rivers then migrate to the seas where they develop larger body sizes, then eventually migrate back to their natal rivers in order to breed (Dodson et al., 2013).
+Resident masu in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives (Morita and Nagasawa).
 
-<!-- Importance of Masu -->
-<!-- 	Vulnerability -->
-<!-- 		Very prone to fishing pressure -->
-<!-- 		Very limited distribution -->
-
-<!-- Life History -->
-<!-- 	What decides if an indivividual is a resident/anadromous? -->
-
-<!-- Genetics of resident vs anadromous -->
-
-<!-- What did we do and find? -->
-<!-- 	Thesis -->
-
-Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in the recent years (CITE).
-In North America, for example,there have been significant decreases indiversity within pacific salmonid species due hydrologic and dam alterations (CITE), and atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams (CITE).
-Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to these dams blocking the fishes' migratory paths (CITE- need a research paper for this number)
-
-Masu salmon (Oncorhynchus masou, otherwise known as cherry salmon) have the most limited distribution among the Pacific salmon (CITE).
-Since they spend a greater portion of their life in freshwater rivers or streams than most other migratory salmon, they are particularly more exposed to fishing and sportfishing pressures (CITE).Their increased vulnerability makes it critical to understand the details that influence and control migratory life history variance in the species.
-Masu in particular (like Oncorhynchus mykiss, steelhead salmon/rainbow trout) are partially migratory, with large resident and anadromous populations (CITE).
-As is typical, anadromous masu are initially born in the rivers then migrate to the seas where they develop larger body sizes, then eventually migrate back to their natal rivers in order to breed (CITE).
-Resident masu in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives (Morita and Nagasawa)
-
-The decision to migrate in masu salmon is thought to be primarily influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced (CITE).
-Consequentially, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous (CITE).
-For example, X et al (YEAR) has shown that the number of resident males have increased over the last X years while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased, which has led to a significant decrease in the number of anadromous individuals overall.
+The decision to migrate in masu salmon is thought to be primarily influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced (Morita et al., 2014; Arostegui & Quinn, 2019).
+Consequently, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous (Morita et al., 2014).
+For example, Morita et al (2014) has shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased, which has led to a significant decrease in the number of anadromous individuals overall.
 
@@
 (2016), on 15,239 single nucleotide polymorphisms (SNPs) found of three SNPs from a gene that was strongly linked with adult migration timing, and that 18 total SNPs that were connected to 60% of the trait variation.
-Despite this, however, growth rates are known to strongly influence anadromy in the species—individuals which grow faster as juveniles are far more likely to stay as residents than otherwise (CITE).
-Indeed, migration in the species seems to be controlled jointly the interplay of genetic and environmental factors:
+Despite this, however, growth rates and body condition are known to influence anadromy in the species—individuals which grow faster as juveniles are more likely to stay as residents than otherwise (Kendall et al., 2014).
+Growth rates are also known to be influenced by large genetic elements that differ between anadromous and resident individuals, implying that genetics may contribute to migration indirectly as well (Miller et al., 2011).
+Taken together, it seems clear that migration in steelhead is  jointly controlled by the interplay of genetic and environmental factors.
 
@@
 To improve our understanding of the mechanisms behind residency and anadromy in the species, we sequenced a set of anadromous and resident masu individuals using high-resolution Restriction Associated Digest (RAD) sequencing.
-However, we did not find any clear indicator of the genetic difference between ouranadromous and resident Masu salmon.
-While this may suggest that there is not that there is a strong genetic basis for Masu salmon migratory life history variation, further studies using a bigger sample size from a larger geographic range would need to be conducted in order to fully investigate the influences of genetics on Masu salmon migratory life cycles.
+However, we did not find any clear indicator of the genetic difference between our anadromous and resident masu salmon.
+While this may suggest that there is not that there is a strong genetic basis for masu salmon migratory life history variation, further studies using a bigger sample size from a larger geographic range would need to be conducted in order to fully investigate the influences of genetics on Masu salmon migratory life cycles.
 
-<!-- Anadromous salmon -> dams causing them to not be able to jump over dams -> is we explore the genetic influences of anadromy and residents in MASU salmon then we can tell where to focus our conservation efforts.
--->
-
-<!-- Parr size was generally larger and maturation was more common at more southern latitudes -->
-
-<!-- Population density and temperature two most important factors affecting the growth of the fish.  -->
-
-<!-- superior development promotes residency (early maturation) -->
-
-<!-- need to do citations -->
 ```{r, include=FALSE, eval=TRUE}
@@
 This could occur if the effect of migratory genes depends on environmental conditions (GxE).
-This phenomena is seen in steelhead, where variance in environmental conditions across the species’ range can determine whether life histories are more genetically or environmentally based (CITE).
+This phenomena is seen in steelhead, where variance in environmental conditions across the species’ range can determine whether life histories are more genetically or environmentally based (Kendall et al., 2014).
 

### 2022-06-01 paper/paper_draft.Rmd
knitting

@@
 ---
-title: "No Detectable Evidence of Genetic Associations Between Anadromous and Resident Masu Salmon Sampled within Hokkaido"
+title: "No Detectable Evidence of Genetic Associations with Anadromous and Resident Masu Salmon Sampled within Hokkaido"
 author: "William Hemstrom"
@@
     fig_height: 6
-bibliography: "Citations.bib"
+bibliography: "MASU_Paper_References.bib"
 csl: "council-of-science-editors-author-date.csl"
@@
 
-Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in recent years (Pringle et al., 2000).
-In North America, for example,there have been significant decreases in diversity within pacific salmonid species due to hydrologic and dam alterations, and atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams (Pringle et al., 2000).
-Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to these dams blocking the fishes' migratory paths (Deinet et al., 2020).
+Due to the history of hydrologic modifications such as the development of reservoirs, dams, and irrigation systems, there has been an apparent decrease in the number of migratory fishes in recent years [@Pringle2000].
+In North America, for example,there have been significant decreases in diversity within pacific salmonid species due to hydrologic and dam alterations, and atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams [@Pringle2000].
+Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to these dams blocking the fishes' migratory paths [@deinet2020living].
 
 Masu salmon (Oncorhynchus masou, otherwise known as cherry salmon) have the most limited distribution among the Pacific salmon.
-Since they spend a greater portion of their life in freshwater rivers or streams than most other migratory salmon, they are particularly exposed to fishing and sportfishing pressures (Miyakoshi et al., 2009).
+Since they spend a greater portion of their life in freshwater rivers or streams than most other migratory salmon, they are particularly exposed to fishing and sportfishing pressures [@Miyakoshi2009].
 Their increased vulnerability makes it critical to understand the details that influence and control migratory life history variance in the species.
-Masu in particular (like steelhead salmon/rainbow trout Oncorhynchus mykiss and other species to a lesser degree) are partially migratory, with large resident and anadromous populations (Morita et al., 2014).
-As is typical, anadromous masu are initially born in the rivers then migrate to the seas where they develop larger body sizes, then eventually migrate back to their natal rivers in order to breed (Dodson et al., 2013).
-Resident masu in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives (Morita and Nagasawa).
+Masu in particular (like steelhead salmon/rainbow trout Oncorhynchus mykiss and other species to a lesser degree) are partially migratory, with large resident and anadromous populations [@Morita2014].
+As is typical, anadromous masu are initially born in the rivers then migrate to the seas where they develop larger body sizes, then eventually migrate back to their natal rivers in order to breed [@Dodson2013].
+Resident masu in contrast sexually mature at a younger age and stay within the river systems throughout their entire lives [@Morita2010].
 
-The decision to migrate in masu salmon is thought to be primarily influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced (Morita et al., 2014; Arostegui & Quinn, 2019).
-Consequently, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous (Morita et al., 2014).
+The decision to migrate in masu salmon is thought to be primarily influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced [@Morita2014; @Arostegui2019].
+Consequently, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous [@Morita2014].
 For example, Morita et al (2014) has shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased, which has led to a significant decrease in the number of anadromous individuals overall.
@@
 It is not surprising, therefore, that a recent study by Hess et al.
-(2016), on 15,239 single nucleotide polymorphisms (SNPs) found of three SNPs from a gene that was strongly linked with adult migration timing, and that 18 total SNPs that were connected to 60% of the trait variation.
-Despite this, however, growth rates and body condition are known to influence anadromy in the species—individuals which grow faster as juveniles are more likely to stay as residents than otherwise (Kendall et al., 2014).
-Growth rates are also known to be influenced by large genetic elements that differ between anadromous and resident individuals, implying that genetics may contribute to migration indirectly as well (Miller et al., 2011).
+[-@Hess2016], on 15,239 single nucleotide polymorphisms (SNPs) found of three SNPs from a gene that was strongly linked with adult migration timing, and that 18 total SNPs that were connected to 60% of the trait variation.
+Despite this, however, growth rates and body condition are known to influence anadromy in the species—individuals which grow faster as juveniles are more likely to stay as residents than otherwise [@Kendall2015].
+Growth rates are also known to be influenced by large genetic elements that differ between anadromous and resident individuals, implying that genetics may contribute to migration indirectly as well [@MILLER2012].
 Taken together, it seems clear that migration in steelhead is  jointly controlled by the interplay of genetic and environmental factors.
@@
 association$p <- pchisq(association$LRT, 1, lower.tail = FALSE)
-plot_manhattan(association, "p", chr = "Chromosome", bp= "Position", log.p = TRUE)
+figure_1 <- plot_manhattan(association, "p", chr = "Chromosome", bp= "Position", log.p = TRUE)
 
@@
 ## Association Testing
-We performed a basic association test on the data using gmmat to correct for the related among individuals and population structure ().
-We then performed a random forest analysis in R-studio using the ranger R package [@Wright2017].
-We used 100,000 trees and used mtry of half the number of snp's.
-We used a random forest to predict whether the samples would be resident or migratory.
-Using the BGLR program, we did a genomic prediction using BayesB distribution to estimate the heritability [@Meuwissen2001].
-We used 1000 burn- ins and iterations=10,000.
+
+We performed a basic association test against migratory phenotype using ANGSD using the first 20 Principal Component Analysis components as covariates to correct for the related among individuals and population structure [@Korneliussen2014].
+To increase our association resolution, we used used genotype likelihoods instead of called gneotypes to conduct our association test, generated in ANGSD using the same parameters described above.
+In order to see if any more complex genetic relationships could predict migratory phenotype, we then performed a random forest analysis in R-studio using the ranger R package [@Wright2017].
+We used 100,000 trees and used mtry of half the number of called SNPs.
+
+<!-- We also used the BGLR program to perform a genomic prediction using BayesB distribution to estimate the heritability [@Meuwissen2001].
+We used 1,000 burn-ins and 10,000 estimation iterations.
 The heritability test was done to determine the degree in which phenotypic variation is due to genetic differences among the salmon populations.
+-->
 
 ## Population Structure and Migratory Relatedness
-In order to detect any underlying population structuring within our stream systems, we used the software package NGSadmix to group samples into *k* clusters for each value of *k* between one and seven [@Skotte693].
-To do so, we first called genotype posterior probailites using ANGSD with the same parameters above in addition to (fill in here).
+<!-- In order to detect any underlying population structuring within our stream systems, we used the software package NGSadmix to group samples into *k* clusters for each value of *k* between one and seven [@Skotte693] using the genotype likelihoods mentioned above.
 To assess variance in clustering within a given *k* value, we ran NGSadmix ten times for each value of *k* and combined the resulting data using CLUMPP [@Jakobsson2007].
-We used the pophelper and snpR packages to handle files and produce plots [@Hemstrom2021; @Francis2017].
+We used the pophelper and snpR packages to handle files and produce plots, respectively [@Hemstrom2021; @Francis2017].
+-->
 
-For comparison, we also conducted a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014].
+We also conducted a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014].
 To look for patterns in relatedness between resident and andromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample the mean for the other population.
-
-<!-- include that we did a one-way anova to determine if either phenotype was more closely related to the mean of the opposite population -->
+We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.
 
@@
 ##Genotyping
-The number of unfiltered reads are `r sort_read_count`.
-After filtering, the number of reads were `r sort_flt_read_count`.
-We called a total of `r number_snps_after_filtering` snps and the number of IBS loci were `r nrow(IBS)`.
-```{r read, include=FALSE}
+
+```{r read, include=FALSE, eval=TRUE}
 library(readr)
 sort_read_count <- read.csv("~/Downloads/sort.read_count.txt", header = FALSE)
-View(sort_read_count)
+# View(sort_read_count)
 sort_read_count <- sum(sort_read_count$V1)
@@
 sort_flt_read_count <- read.csv("~/Downloads/sort.flt.read_count.txt", header = FALSE)
-View(sort_flt_read_count)
+# View(sort_flt_read_count)
 sort_flt_read_count <- sum(sort_flt_read_count$V1)
@@
 
-flt read = `r sort_flt_read_count` 
+The number of unfiltered reads are `r prettyNum(sort_read_count,big.mark=",",scientific=FALSE)`.
+After filtering, the number of reads were `r prettyNum(sort_flt_read_count,big.mark=",",scientific=FALSE)`.
+We called a total of `r prettyNum(number_snps_after_filtering,big.mark=",",scientific=FALSE)` SNPs and `r prettyNum(nrow(IBS),big.mark=",",scientific=FALSE)` IBS loci.
+<!-- flt read = `r sort_flt_read_count`  -->
 <!-- flt read = 90449126 -->
-unfiltered reads = `r sort_read_count` 
+<!-- unfiltered reads = `r sort_read_count`  -->
 <!-- unfiltered reads = 163200129 -->
-number of snps = `r number_snps_after_filtering`
-Number of snps= 71175?
-number of IBS loci = `r nrow(IBS)`
+<!-- number of snps = `r number_snps_after_filtering` -->
+<!-- Number of snps= 71175? -->
+<!-- number of IBS loci = `r nrow(IBS)` -->
 
 ##  Diversity Stats
-The mean Ho values are relatively similar between the anadromous and residents individuals and between the CH and TK populations.
-The fst values between the two populations and between anadromous and resident individuals are low, which indicated that there is not much genetic diversity between the two populations and two phenotypes.
-There are also no private alleles between the two populations, which also indicated low genetic diversity between the two populations of salmon.
+The mean $H_{o}$ values are relatively similar between the anadromous and residents individuals and between the CH and TK populations (see Table 1).
+The $F_{ST}$ values between the two populations and between anadromous and resident individuals were very low, which indicated that there is not much genetic diversity between the two populations and two phenotypes.
+There are also no private alleles between the two populations, which also indicated low genetic differentiation between the two populations of salmon.
 
 ## Association
-The random forest predicted 25 out of 52 anadromous individuals correctly and 17 out of 44 resident individuals correctly for a total prediction accuracy of 38.6%.
+We did not find any strong associations with migratory life history status at any loci (Figure 1), and the random forest only successfully predicted 25 out of 52 anadromous individuals correctly and 17 out of 44 resident individuals correctly for a total prediction accuracy of 38.6%.
 
 ## Influence of Anadromy on Population Structure
-We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.
-We found that the anadromous individuals clustered signficiantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419).
-This indicates that the anadromous individuals within one population are more closely related to the individuals in the other population.
+ We found that the anadromous individuals clustered significantly closer to the cluster center for individuals from the other population (one-way ANOVA, p-value = 0.0419, see Figure 2).
+This indicates that the anadromous individuals within one population were more closely related to the individuals in the other population than were residents.
 <!-- The p value (phenR) is 0.03981775. The p value for the phenotypes is 0.0419.  -->
@@
 This could occur if the effect of migratory genes depends on environmental conditions (GxE).
-This phenomena is seen in steelhead, where variance in environmental conditions across the species’ range can determine whether life histories are more genetically or environmentally based (Kendall et al., 2014).
+This phenomena is seen in steelhead, where variance in environmental conditions across the species’ range can determine whether life histories are more genetically or environmentally based [@Kendall2015].
 
@@
 
-<!-- control shift c --> 
-[@Gowen2014; @Ali2016]
-Gowen et al.[-@Gowen2014]
+# References:
+
+<div id="refs"></div>
+\newpage
+
+# Figures:
+
+```{r Figure_1_print,echo=FALSE,warning=FALSE,fig.cap="Figure 1: Association across the genome with migratory status showing that no loci were found that were significantly associated with the given traits (anadromy and residency)",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
+figure_1$plot
+
+```
+
+```{r Figure_2_print,echo=FALSE,warning=FALSE,fig.cap="Figure 2: PCA plot showing the relatedness between anadromous and resident individuals in TK and CH populations",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
+p1
+
+```

### 2022-06-01 paper/paper_draft.Rmd
knit

@@
 library(readr)
-sort_read_count <- read.csv("~/Downloads/sort.read_count.txt", header = FALSE)
+sort_read_count <- read.csv("../sort.read_count.txt", header = FALSE)
 # View(sort_read_count)
@@
 
-sort_flt_read_count <- read.csv("~/Downloads/sort.flt.read_count.txt", header = FALSE)
+sort_flt_read_count <- read.csv("../sort.flt.read_count.txt", header = FALSE)
 # View(sort_flt_read_count)

### 2022-06-01 paper/paper_draft.Rmd
knit

@@
 gp <- readRDS("gp.RDS")
+number_snps_after_filtering <- nsnps(dat)
 
@@
 ##  Diversity Stats
-We calculated the observed heterozygosity ($H_{o}$), the average number of pairwise differences ($\pi$), number of private alleles, and fixation index ($F_{st}$) between populations using the snpR R package [@Hohenlohe2010; @Wier1984; @Hemstrom2021].
+We calculated the observed heterozygosity ($H_{o}$), the average number of pairwise differences ($\pi$), number of private alleles, and fixation index ($F_{st}$) between populations using the snpR R package [@Hohenlohe2010; @Weir1984; @Hemstrom2021].
 

### 2022-06-01 paper/paper_draft.Rmd
hash

@@
 
-#Abstract
+# Abstract
+
+Within Masu salmon (Oncorhynchus masou), individuals can exhibit one of two life histories: resident individuals which stay in streams for their entire life span or anadromous individuals which move to the ocean to mature before returning to their natal streams to breed.
+Current research has suggested that the expression of either life history is mainly influenced by environmental factors such as temperature, which influence critical traits like salmon's body size, growth rate, lipid stores.
+Here, we use high-resolution reduced-representation genome sequencing to show that there are no significant genetic associations with migratory decision making in masu salmon in two streams in Hokkaido, Japan.
+Further research with a larger sample size from more parts of the masu range is needed in order to fully assess potential genetic influences on anadromy and residency in the species.
+
+# Introduction
 
@@
 
-##Genotyping
+## Genotyping
 
@@
 
-```{r Figure_1_print,echo=FALSE,warning=FALSE,fig.cap="Figure 1: Association across the genome with migratory status showing that no loci were found that were significantly associated with the given traits (anadromy and residency)",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
+```{r Figure_1_print,echo=FALSE,warning=FALSE, include=TRUE,fig.cap="Figure 1: Association across the genome with migratory status showing that no loci were found that were significantly associated with the given traits (anadromy and residency)",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
 figure_1$plot
@@
 
-```{r Figure_2_print,echo=FALSE,warning=FALSE,fig.cap="Figure 2: PCA plot showing the relatedness between anadromous and resident individuals in TK and CH populations",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
+```{r Figure_2_print,echo=FALSE,warning=FALSE,include=TRUE,fig.cap="Figure 2: PCA plot showing the relatedness between anadromous and resident individuals in TK and CH populations",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
 p1

### 2022-06-01 paper/paper_draft.Rmd
IBS

@@
 IBS <- data.table::fread("../angsdput.covMat")
+IBS.loci <- data.table::fread("../data/angsdput.ibs.gz")
 nrow(IBS)
@@
 After filtering, the number of reads were `r prettyNum(sort_flt_read_count,big.mark=",",scientific=FALSE)`.
-We called a total of `r prettyNum(number_snps_after_filtering,big.mark=",",scientific=FALSE)` SNPs and `r prettyNum(nrow(IBS),big.mark=",",scientific=FALSE)` IBS loci.
+We called a total of `r prettyNum(number_snps_after_filtering,big.mark=",",scientific=FALSE)` SNPs and `r prettyNum(nrow(IBS.loci),big.mark=",",scientific=FALSE)` IBS loci.
 <!-- flt read = `r sort_flt_read_count`  -->
@@
 
-We did not find any significant genetic associations with migratory phenotype (see Figure X).
+We did not find any significant genetic associations with migratory phenotype (see Figure 1).
 This indicates that in the streams we examined, there may not be any loci in the genome that strongly contribute to migratory life history decisions.

### 2022-06-01 paper/paper_draft.Rmd
added the fixed, knitted paper draft since it wasn't knitting correctly for Audrey

@@
 ```{r, eval=TRUE, include=FALSE}
-mat <- read.table("../angsdput.covMat")
+# mat <- read.table("../angsdput.covMat")
 IBS <- data.table::fread("../angsdput.covMat")
@@
 
-```{r Figure_1_print,echo=FALSE,warning=FALSE, include=TRUE,fig.cap="Figure 1: Association across the genome with migratory status showing that no loci were found that were significantly associated with the given traits (anadromy and residency)",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
+```{r Figure_1_print,echo=FALSE,warning=FALSE,eval=TRUE,include=TRUE,fig.cap="Figure 1: Association across the genome with migratory status showing that no loci were found that were significantly associated with the given traits (anadromy and residency)",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
 figure_1$plot
@@
 
-```{r Figure_2_print,echo=FALSE,warning=FALSE,include=TRUE,fig.cap="Figure 2: PCA plot showing the relatedness between anadromous and resident individuals in TK and CH populations",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
+```{r Figure_2_print,echo=FALSE,warning=FALSE,eval=TRUE,include=TRUE,fig.cap="Figure 2: PCA plot showing the relatedness between anadromous and resident individuals in TK and CH populations",fig.width=7, fig.height=4, fig.align='center', fig.cap.style="Image Caption"}
 p1

### 2022-08-31 paper/paper_draft.Rmd
updated draft

@@
 In North America, for example,there have been significant decreases in diversity within pacific salmonid species due to hydrologic and dam alterations, and atlantic salmon and anadromous species of sturgeons are also impacted negatively because of a loss of portions to their habitats upstream of dams [@Pringle2000].
-Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to these dams blocking the fishes' migratory paths [@deinet2020living].
+Overall, the diversity within migratory fish species have largely declined by an average of 70% between 1970 and 2016 due to dams and other factors  [@deinet2020living].
 
@@
 The decision to migrate in masu salmon is thought to be primarily influenced by environmental factors such as temperature alongside other factors such as the salmon's body size, growth rate, lipid stores, and is heavily sex influenced [@Morita2014; @Arostegui2019].
+Migration and residency in the closely related steelhead salmon (Oncorhynchus mykiss) are controlled by similar factors.
+For example, resident steelhead tend to be larger within the first few months of life when compared to anadromous individuals (Morita et al., 2014).
+Of course, differences in environmental factors can influence growth rates and therefore also migration in steelhead.
+For example, during poor nutrient conditions, a higher proportion of individuals than expected from genetics may be anadromous; whereas during flooding or nutrient rich periods, a higher proportion of resident individuals may be observed [@Kelson2020].
+Regardless of environmental conditions, however, those individuals within a population which grow faster as juveniles are more likely to stay as residents than otherwise, even under similar conditions [@Kendall2015].
+
+Migration comes with a tradeoff.
+In steelhead, migratory juveniles tend to have low survival rates to adulthood, which is  compensated for by the high fecundity of anadromous females that can significantly exceed that of residents [@Pearse2019].
+This pattern likely holds true for masu as well (Morita et al., 2014).
+This trade-off is not as pronounced in male steelhead, which can mature as early as freshwater residents, avoiding the high mortality associated with migration [@Pearse2019].
+This results in greater frequency of anadromy in females; a trend also seen in masu salmon (Morita et al., 2014).
 Consequently, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous [@Morita2014].
-For example, Morita et al (2014) has shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as the temperatures have increased, which has led to a significant decrease in the number of anadromous individuals overall.
+For example, Morita et al (2014) has shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as temperatures have increased over time, leading to a significant decrease in the overall number of anadromous individuals.
 
-In contrast, the genetic factors that contribute to migration in masu salmon have not been well studied.
+In contrast to environmental factors, the genetic factors that contribute to migration in masu salmon have not been well studied.
 Despite the prevalence of environmental control of migration in masu, it is not unreasonable to expect genetic elements to play some role in governing migratory decisions in the species.
@@
 In 1944, Naeve conducted common garden and transplant experiments with sympatric rainbow trout and steelhead and found that migratory status was heritable after observing that rainbow trout offspring were considerably more likely to be resident themselves, whereas crosses between anadromous individuals resulted in the most anadromous smolts.
-It is not surprising, therefore, that a recent study by Hess et al.
-[-@Hess2016], on 15,239 single nucleotide polymorphisms (SNPs) found of three SNPs from a gene that was strongly linked with adult migration timing, and that 18 total SNPs that were connected to 60% of the trait variation.
-Despite this, however, growth rates and body condition are known to influence anadromy in the species—individuals which grow faster as juveniles are more likely to stay as residents than otherwise [@Kendall2015].
-Growth rates are also known to be influenced by large genetic elements that differ between anadromous and resident individuals, implying that genetics may contribute to migration indirectly as well [@MILLER2012].
+Indeed, recent studies have found that  the decision to migrate in steelhead is strongly linked to a large genomic inversion on chromosome Omy5 throughout their geographic range which seems to maintain a set of linked genes that controls migratory behavior [@Arostegui2019; @Kelson2020; @Pearse2014).
+The genetic basis for migration is also likely indirectly mediated by physiological characteristics such as growth rate, which are also known to be influenced by the same large genetic elements that differ between anadromous and resident individuals [@Miller2021].
 Taken together, it seems clear that migration in steelhead is jointly controlled by the interplay of genetic and environmental factors.
@@
 library(snpR)
-# dat<- import.snpR.data(dat[,-c(1:2)], snp.meta = snp.meta, sample.meta = meta)
-# dat<- filter_snps(dat, maf = 0.05, hwe = 0.000001)
-# number_snps_after_filtering <- nsnps(dat)
-# 
-# #calculating genetic diversity
-# dat <- calc_ho(dat, c("pop", "phen", "pop.phen"))
-# get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="ho") $weighted.means
-# get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
-# dat <- calc_fis(dat, c("pop", "phen"))
-# dat <- calc_pi(dat, c("pop", "phen"))
-# get.snpR.stats(dat, c("pop"), stats="pi") $weighted.means
-# dat <- calc_private(dat, c("phen", "pop","pop.phen"))
-# get.snpR.stats(dat, c("pop"), stats="private") $weighted.means
-# get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
-# dat <- calc_pairwise_fst(dat, c("phen", "pop","pop.phen"), boot = 100, boot_par = 2)
-# get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="fst") $weighted.means
-# get.snpR.stats(dat, c("pop", "phen"), stats= c("ho", "pi", "fst")) $weighted.means
+dat<- import.snpR.data(dat[,-c(1:2)], snp.meta = snp.meta, sample.meta = meta)
+dat<- filter_snps(dat, maf = 0.05, hwe = 0.000001)
+number_snps_after_filtering <- nsnps(dat)
+
+#calculating genetic diversity
+#dat <- calc_ho(dat, c("pop", "phen", "pop.phen"))
+#get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="ho") $weighted.means
+#get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
+#dat <- calc_fis(dat, c("pop", "phen"))
+#dat <- calc_pi(dat, c("pop", "phen"))
+#get.snpR.stats(dat, c("pop"), stats="pi") $weighted.means
+#dat <- calc_private(dat, c("phen", "pop","pop.phen"))
+#get.snpR.stats(dat, c("pop"), stats="private") $weighted.means
+#get.snpR.stats(dat, c("pop", "phen", "pop.phen"))
+#dat <- calc_pairwise_fst(dat, c("phen", "pop","pop.phen"), boot = 100, boot_par = 2)
+#get.snpR.stats(dat, c("phen", "pop","pop.phen"), stats="fst") $weighted.means
+#get.snpR.stats(dat, c("pop", "phen"), stats= c("ho", "pi", "fst")) $weighted.means
 # 
@@
 We also conducted a Principle Component Analysis chart (PCA) using the ANGSD-IBS method [@Korneliussen2014].
-To look for patterns in relatedness between resident and andromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample the mean for the other population.
+To look for patterns in relatedness between resident and anadromous individuals, we calculated the mean PC1 and PC2 coordinates for each stream, then calculated the distance between each individual sample the mean for the other population.
 We fit a simple linear regression with the distance to the other population cluster as the response variable and the migratory phenotype as the explanatory variable.

### 2022-08-31 paper/paper_draft.Rmd
MASU draft

@@
 bibliography: "MASU_Paper_References.bib"
-csl: "council-of-science-editors-author-date.csl"
+csl: "molecular-ecology.csl"
 header-includes:
@@
 
-Masu salmon (Oncorhynchus masou, otherwise known as cherry salmon) have the most limited distribution among the Pacific salmon.
+Masu salmon (Oncorhynchus masou, otherwise known as cherry salmon) have the most limited distribution among the Pacific salmon [@augerot2005atlas].
 Since they spend a greater portion of their life in freshwater rivers or streams than most other migratory salmon, they are particularly exposed to fishing and sportfishing pressures [@Miyakoshi2009].
@@
 Migration and residency in the closely related steelhead salmon (Oncorhynchus mykiss) are controlled by similar factors.
-For example, resident steelhead tend to be larger within the first few months of life when compared to anadromous individuals (Morita et al., 2014).
+For example, resident steelhead tend to be larger within the first few months of life when compared to anadromous individuals [@Morita2014].
 Of course, differences in environmental factors can influence growth rates and therefore also migration in steelhead.
@@
 In steelhead, migratory juveniles tend to have low survival rates to adulthood, which is  compensated for by the high fecundity of anadromous females that can significantly exceed that of residents [@Pearse2019].
-This pattern likely holds true for masu as well (Morita et al., 2014).
+This pattern likely holds true for masu as well [@Morita2014].
 This trade-off is not as pronounced in male steelhead, which can mature as early as freshwater residents, avoiding the high mortality associated with migration [@Pearse2019].
-This results in greater frequency of anadromy in females; a trend also seen in masu salmon (Morita et al., 2014).
+This results in greater frequency of anadromy in females; a trend also seen in masu salmon [@Morita2014].
 Consequently, changes in environmental factors, such as those caused by climate change, have been known to impact the decision of the masu to be either resident or anadromous [@Morita2014].
-For example, Morita et al (2014) has shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as temperatures have increased over time, leading to a significant decrease in the overall number of anadromous individuals.
+For example, Morita et al [-@Morita2014] has shown that the number of resident males have increased while the proportion of anadromous males and the proportion of delayed migrants within both females and males have decreased as temperatures have increased over time, leading to a significant decrease in the overall number of anadromous individuals.
 
@@
 For example, within steelhead, there is strong evidence suggesting both an environmental and genetic basis for migratory life history.
-In 1944, Naeve conducted common garden and transplant experiments with sympatric rainbow trout and steelhead and found that migratory status was heritable after observing that rainbow trout offspring were considerably more likely to be resident themselves, whereas crosses between anadromous individuals resulted in the most anadromous smolts.
-Indeed, recent studies have found that  the decision to migrate in steelhead is strongly linked to a large genomic inversion on chromosome Omy5 throughout their geographic range which seems to maintain a set of linked genes that controls migratory behavior [@Arostegui2019; @Kelson2020; @Pearse2014).
-The genetic basis for migration is also likely indirectly mediated by physiological characteristics such as growth rate, which are also known to be influenced by the same large genetic elements that differ between anadromous and resident individuals [@Miller2021].
+In 1944, Naeve conducted common garden and transplant experiments with sympatric rainbow trout and steelhead and found that migratory status was heritable after observing that rainbow trout offspring were considerably more likely to be resident themselves, whereas crosses between anadromous individuals resulted in the most anadromous smolts [double check paper that this was in].
+Indeed, recent studies have found that  the decision to migrate in steelhead is strongly linked to a large genomic inversion on chromosome Omy5 throughout their geographic range which seems to maintain a set of linked genes that controls migratory behavior [@Arostegui2019; @Kelson2020; @Pearse2014].
+The genetic basis for migration is also likely indirectly mediated by physiological characteristics such as growth rate, which are also known to be influenced by the same large genetic elements that differ between anadromous and resident individuals [@MILLER2012].
 Taken together, it seems clear that migration in steelhead is jointly controlled by the interplay of genetic and environmental factors.
